Mediodorsal thalamic nucleus mediates resistance to ethanol through Cav3.1 T-type Ca2+ regulation of neural activity

Version 1

Abstract

Abstract

Thalamocortical activity is known to orchestrate sensory gating and consciousness switching. The precise thalamic regions involved, or the firing patterns related to the unconsciousness remain unclear. Interestingly, the thalamically highly-expressed T-type calcium currents have been considered as a candidate for the ionic mechanism for the generation of thalamic-driven change in conscious state. Here, we tested the hypothesis that Cav3.1 T-type channels in the mediodorsal thalamic nucleus (MD) might control neuronal firing during unconsciousness using Cav3.1 T-type channel knock-out (KO) and knock-down (KD) mice under natural sleep and ethanol-induced unconsciousness. During natural sleep, the MD neurons in KO mice showed general characteristics of sustained firing across sleep stages. We found that KO and MD-specific KD mice showed enhanced resistance to ethanol. During ethanol-induced unconscious state, wild-type (WT) MD neurons showed a significant reduction in neuronal firing from baseline with increased burst firing, whereas Cav3.1 KO neurons showed well sustained neural firing, within the level of wakefulness, and no burst firing. Further, 20 Hz optogenetic and electrical activation of MD neurons mimicked the ethanol resistance behavior in WT mice. These results support that the maintenance of MD neural firing at a wakeful level is sufficient to cause resistance to the ethanol hypnosis in WT mice. This work has important implications for the design of treatments for consciousness disorders using thalamic stimulation of deeper nuclei including the targeting of the mediodorsal thalamic nucleus.

Introduction

Drug-induced unconsciousness can be achieved using numerous types of anesthetics with varying modes of action 1,2. Ethanol, one of the most frequently abused drugs in human society, can induce sleep-like loss of consciousness at high doses 3. While possible neuropharmacological and neural correlates of ethanol sedation have been proposed using in vitro and in vivo methods 4–7, recent studies have highlighted the slowing of thalamocortical-driven rhythms as a potent marker of unconsciousness 8,9. However, the region and the mechanism linked to the thalamic modulation during ethanol induced unconsciousness remains poorly understood.

Physiological correlates of thalamocortical rhythmic activities and consciousness state of the brain are known 10–13. T-type calcium channels are known generators of thalamocortical rhythms, through the modulation of cell excitability and rebound burst firing 14,15. During sleep, the transition from wakefulness to unconsciousness is associated with membrane hyperpolarization of thalamic neurons 12,16. Similarly, It has been shown that in ethanol sedation, as in natural sleep or absence seizure, the loss of consciousness is characterized by a switch from tonic to burst firing in thalamic neurons, which involves GABAergic inhibition-driven de-inactivation of Cav3.1 T-type channels resulting in slow oscillatory response of the thalamocortical network 7,17–19. Acute intoxication at high doses of ethanol 20,21 induces both slow oscillations in the delta-theta frequency range and a loss of righting reflex (LORR), the loss of reflex to up-right itself from a supine position, a classical proxy to assess the loss of consciousness. It has been shown that mice lacking global or thalamic Cav3.1 showed altered slow oscillations and sleep architecture 22,23; delayed sleep induction under several anesthetics (i.e. isoflurane, halothane, sevoflurane and pentobarbital) 24; and increased resistance to drug-induced absence seizure 13. Notably, the absence or blockade of Cav3.1 resulted in an increased preference for ethanol consumption and novelty-seeking behavior 20,21. In the current studies we tried to explore the possibility of a pivotal role of Cav3.1-mediated T-currents during ethanol-induced sleep or unconsciousness.

The thalamus is one of the major regions expressing Cav3.1 T-type calcium channels 25 and holds a central role in information-transmission and integration 26. In vitro and in vivo studies using genetically modified mice have revealed that Cav3.1 T-type channels play a key role in the genesis of thalamocortical rhythms, such as 3Hz spike-and-wave discharges (SWD), a signature of absence seizures 13,27 and delta waves 16,28,29. Previous investigations on thalamic control of consciousness revealed that nuclei within the dorsal medial thalamus (dMT) hold an important modulatory function in the interaction of attention and arousal 9,30. Particularly, centromedian (CM) thalamic nucleus, and not ventrobasal nucleus (VB), showed rapid shifts in local field potential (LFP) preceding brain state transitions such as NREM and propofol-induced anesthesia 8. The centrolateral (CL) thalamic nucleus has been implicated in the modulation of arousal, behavior arrest 31, and improvement of level of consciousness during seizures 32 and paraventricular thalamic nucleus showed critical involvement in wake/sleep cycle regulation 33. The MD, a sub nucleus of dMT, on the other hand, has only recently been implicated in disorders of consciousness 34 and ketamine/ethanol-induced loss of consciousness 35 through the alteration of thalamo-cortical functional connectivity. However, several key questions still remain to be answered: 1) Is there a specific role for MD Cav3.1 T-type calcium channels in the control of ethanol-induced loss of consciousness? 2) Whether Cav3.1 T-type calcium channel driven neuronal firing pattern has any role in the control of consciousness?

In this study, we identified that KO and MD specific silencing of Cav3.1 T-type calcium channels result in increased ethanol resistance in mice. Using single unit recordings, we compared MD activity of WT and KO mice while the mice transitioned from conscious to unconscious state and found that the KO mice showed more sustained MD activity whereas the WT mice showed clearly reduced MD activity. Furthermore and consistently with their resistant phenotype, KO mice showed sustained MD firing, well within the wakefulness level, under ethanol consumption. Finally, we demonstrate that both the optogenetic and electrical stimulations in MD that were mimicking the sustained firing pattern of knock-out mice were sufficient to induce the increased ethanol resistance in WT mice. These results reveal a novel function of MD in ethanol-induced unconsciousness and its underlying neural mechanism.

Results

To understand the role of T-type Ca2+ channels in modulating the consciousness level, we compared the ethanol resistance between WT and Cav3.1−/− KO littermates. We used the forced walking task (FWT; Fig. 1-A), an analog to the LORR assay, which enables a continuous and high-temporal resolution assessment of the loss of movement36 (LOM). Moreover, the FWT objectively measures the latency to and duration of the first LOM, but also the total time spent in LOM using automatized analysis of video confirmed by electromyograms (EMGs) or accelerometer recordings (Fig. 1-B and S1; see Methods). The continuously running treadmill (6 cm/s) ensures a normalized behavior within and between animals before injection (i.e., baseline forced walking) and allows for reduced intervention from experimenters for the monitoring of both electrophysiological (Fig. S1, upper panel) and analyzed behavioral (Fig. S1, lower panel).

Mice lacking Cav3.1 showed increased ethanol resistance on the forced walking task.

(A) The schematic of the FWT setup. Mice are habituated and trained on a constantly moving treadmill (6 cm/sec). Following a baseline walking recording (∼10 min), the mouse is carefully picked up and injected with ethanol (i.p.). Once placed back on the treadmill, the loss of consciousness is evaluated using normalized moving index using either video analysis (differential pixel motion), on-head accelerometer-based motion, or neck electromyograms over a period of 60 min. (B) Representative EEG (parietal) and xyz accelerometer (Acc) traces for walking and LOM in a Cav3.1 (+/+) mouse. (C) Quantification for the latency to first LOM (fLOM; i.e. delay between I.P. Injection and loss of motion; C1), the duration of the fLOM (C2) and the total time spent in LOM state (C3) over a recording duration of 60 min following 3.0 g/Kg I.P injection of ethanol in Cav3.1 (+/+) and Cav3.1 (-/-) mice; data is represented as boxplot. * is for p<0.05, ** is for p<0.01 and *** is for p<0.001. (D) Representative normalized motor activity over time for Cav3.1(+/+) (D1) and Cav3.1 (-/-) (D2) mice post I.P. injection of 3.0 g/Kg. Blue and red boxes above the graph indicate the state interpretation for walking and LOM, respectively.

The lacking Cav3.1 increases ethanol resistance in mice

Mice lacking Cav3.1 exhibited delayed anesthetic induction 13,24 and impairment in maintenance of low conscious level 22,23 as well as increased ethanol preference 21,35. We tested the sensitivity of Cav3.1 null mutant mice for various acute hypnotic doses of ethanol.

We confirmed that the lack of Cav3.1 resulted in a more delayed and fragmented LOM (Fig. 1-C1 and C2), and a reduction in the total time spent in LOM compared to Cav3.1 WT mice (Fig. 1-C3). We observed that Cav3.1 null mutant mice showed increased latency to and decreased duration of the first episode of loss of motion (fLOM) for ethanol injection doses of 2.0, 3.0 and 4.0 g/Kg (Fig. S2). The total time spent in LOM during one hour of recording was also significantly reduced (Fig. S2-A3 and -B3) compared with WT mice. Two-way analysis of variance (ANOVA) showed a significant effect for the main factor genotype and dose for the latency to and duration of fLOM, and total time in LOM (Table S1), indicating a dose dependency in both wild and mutant mice. In particular, an I.P. Injection of 3.0 g/Kg induced a significant difference between wild and mutant mice in latency to (t(16) =-4.1965, p = 0.002; Student t-test) and duration of fLOM (t(16) =2.3908, p = 0.0294; Student t-test), and total time spent in LOM (t(16) =3.9065, p = 0.0012; Student t-test).

These results indicate that ethanol induces a dose-dependent sedative effect on mice and Cav3.1 mutant mice had an increased resistance to ethanol sedation compared to WT mice.

Cav3.1 silencing in the MD, but not VB, increased ethanol resistance in mice

A great majority of Cav3.1 expression is found in the thalamic region 25 and was shown to specifically correlate with the modulation of thalamocortical-related rhythms and stability of sleep level 22,23. To identify a possible role of the thalamic region in ethanol resistance, we knocked down the expression of Cav3.1 in the MD and a ventral basal nucleus (VB) region, two regions possibly involved in a thalamic control of consciousness 34,35,37, using a lentivirus (LV)-mediated short hairpin (shRNA) delivery.

We found that compared to shControl injected mice, shCav3.1 knock-down of MD resulted in an increased latency to (Fig. 2-A1; t(27) = −3.0045, p = 0.0057; Student t-test) and duration of (Fig. 2-A2; t(27) =2.1448, p = 0.0411; Student t-test) fLOM, and total time spent in LOM (Fig. 2-A3; t(27) =2.6641, p = 0.0128, two-tailed test) for 3.0g/Kg I.P. Injection of ethanol. However, we found that compared to shControl-injected mice, Cav3.1 KD of VB did not change the latency to (t(8) =-1.0093 p = 0.3423, two-tailed test), duration of (t(8) =-0.0983, p = 0.9241, two-tailed test) fLOM and total time spent in LOM (t(8) =-0.6317, p = 0.5452, two-tailed test) for the same 3.0g/Kg I.P. Injection of ethanol (Fig. S3). Representative traces of mice activity showed that mice with Cav3.1 KD in MD (Fig. S4-A) had a more delayed and fragmented early period of LOM compared to MD LV-shControl and VB injected mice, as in mutant mice.

Cav3.1 knock-down in MD increased ethanol resistance in mice.

(A) Quantification for the Latency to first LOM (fLOM; A1), the duration of the first LOM (A2) and the total time spent in LOM state (A3) over a recording duration of 60 min post I.P. injection of ethanol (3.0g/Kg) in lentivirus-shControl and shCav3.1 knock-down mice for MD; data is represented as boxplot. * and ** indicates p<0.05 and p<0.01, respectively. (B) Representative brain coronal section stained using Cav3.1 antibody (B1) showing the endogenous thalamic expression of Cav3.1 in the MD of lentivirus(LV)-shControl injected mice; Cav3.1 and GFP merging (B2) and higher magnification of the white dashed square in B2 (B3). Scale bars in (B2) and (B3) indicate 500 μm and 100 μm, respectively. (C) Representative brain coronal section showing the reduced Cav3.1 expression in the MD of LV-shCav3.1 injected mice (C1); Cav3.1 and GFP merging (C2) and higher magnification of the white dashed square in C2 (C3). Scale bars in (C2) and (C3) indicate 500 μm and 100 μm, respectively.

During the open field test, Cav3.1 null mutant mice showed significantly increased locomotor activity compared to WT mice as shown by total distance moved (Fig. S4-C; analysis of variance: GROUP F(3) = 8.45, p = 0.0004; Cav3.1+/+ vs Cav3.1-/-: p = 0.0001). The mice with MD-specific Cav3.1 KD, however, did not show any significant difference in total distance moved compared to shControl-injected control mice (MD LV-shControl vs MD LV-shCav3.1: p = 0.868; Holm-Sidak correction), indicating that the ethanol resistance in MD Cav3.1 KD mice was not attributed by hyperlocomotion observed in Cav3.1 KO mice.

Lack of Ca3.1 in MD neurons removes thalamic burst in NREM sleep

Thalamic neurons are known to follow a state-dependent activity 38; however, the nature of this state-dependent activity has not been studied for the MD. In order to understand the relationship between MD neuron firing and level of consciousness, we investigated the association between the neural activity in MD and brain states at different levels of consciousness (Fig. 3 and S5).

Lack of Cav3.1 removed burst firing, and reduces neural activity and its variability in MD across natural conscious and unconscious states.

(A) Mice were implanted unilaterally with 4 tetrode wires to record single unit activity in the MD while in homecage (wake, sleep: NREM, REM) and forced walking task under ethanol (walk). (B - Left panel) Scatter distribution of spike width vs bursting index of MD regular spiking (RS, round shape) and narrow spiking (NS, triangle shape) neurons. Cav3.1 (+/+) and Cav3.1 (-/-) neurons are marked as filled and empty shapes, respectively. The histogram of the pooled Cav3.1 (+/+) and Cav3.1 (-/-) distribution is projected on each axis. (B - Right panel) Representative auto-cross correlograms of a RS neuron showing the presence and absence of fast spiking interval (burst firing) in Cav3.1 wild-type and mutant mice, respectively. (C) Boxplots of Cav3.1 wild and mutant burst firing rate (spikes/sec; burst spikes-only averaged over a state duration) (C1) and burst event rate (#/min; number of burst event averaged over a state duration; see burst definition in methods section) (C2) in RS neurons of the MD during NREM sleep, a stage known for the presence of bursting firing mode in thalamic neurons, for Cav3.1 (+/+) and Cav3.1 (-/-) mice; The inset numbers in C1 indicate the number of neurons showing burst firing (more than 1 event in 10 min) over the total number of single neurons identified. (D) Boxplots of Cav3.1 wild and mutant tonic firing rate (spikes/sec) in RS neurons of the MD during walking (FWT), wake (home cage), REM and NREM sleep (home cage) for Cav3.1 (+/+) and Cav3.1 (-/-) mice. Group and brain state effect and interaction were assessed using a two-way repeated-measure ANOVA. For post-hoc, two samples ranksum test comparison *, ** and *** indicate p-value lower than 0.05, 0.01 and 0.001, respectively. For two sample Levene’s test for homoscedasticity #, ## and ### indicate p-value lower than 0.05, 0.01 and 0.001, respectively. Pearson ranksum correlations between brain state and total firing for Cav3.1(+/+) and Cav3.1(-/-) is indicated above boxplots.

We observed two major populations of neuronal spike waveform present in MD single unit recordings of Cav3.1 (+/+) mice: 1) a majority of regular spiking (RS) cells characterized by wide spike waveform (36/39 neurons, 92.3%), i.e. spike-to-valley width >250 μsec (Fig. 3B) and high bursting propensity; 2) a minority of narrow spiking (NS) cells showing short spike-to-valley width, i.e. <250 μsec, lower bursting characteristics (Fig. 3 and S6-A3 and -A4) and fast-paced tonic firing (10-50Hz; data not shown). The RS and NS neurons were found in MD of WT and mutant mice; however, MD RS mutant neurons showed an absence of short inter-spike intervals (ISI, i.e. indicative of the absence of burst) in auto cross-correlogram (Fig. S6) and a clear reduction in total bursting (Fig. 3-B; ratio of spikes count <10 ms and >50 ms based on auto-cross-correlogram). Since RS cells have the profile of the major population of MD, i.e. excitatory neurons, we focused on the analysis of RS neurons mainly in the remainder of this study.

During the deep sleep state non-rapid eye movement (NREM), thalamic neural firing is known to switch from tonic to burst firing 39. We found that a lack of Cav3.1 T-type calcium channels resulted in a near absence of burst (see Methods for definition) in mutant mice (Fig. 3-C1 and C2; 4/44 bursting neurons; Z(77) = 7.20, p < 0.0001, Ranksum test) compared to WT mice (34/34 bursting neurons; 5.76 ± 5.51 burst events/min).

Lack of Ca3.1 reduces neuronal activity across all brain states in MD

In addition, we observed that the mutant mice showed a significant lower total firing rate (main factor group: F(1,186) = 16.5, p = 0.0001; interaction group x brain state: F(3,186) = 4.72, p = 0.0034) and a reduced variability (p <0.0001 for all brain states; Levene’s test) in most brain states compared to the wild type mice, indicating that the lack of Cav3.1 t-type channels results in an overall reduction in neural activity in RS neurons.

RS neurons of MD in Cav3.1 wild type and mutant mice showed a significant change in overall firing across walking, waking (homecage), NREM and REM sleep states as shown by a repeated measure anova (Fig. 3-D; main factor brain state: F(3,186) = 104.96, p <0.0001). In addition, Cav3.1 wild type (R = −0.534, p = 1.6e-10, Spearman ranked correlation) and mutant (R = −0.689, p = 6.8e-20, Spearman ranked correlation) showed a significant negative correlation between neural firing and brain state. Assuming an ordering from higher to lower state of consciousness, these results indicate that MD firing is associated with level of consciousness independently from the Cav3.1 T-type channels in wild type and mutant mice. Importantly, this result indicates that Cav3.1 t-type calcium channels are critical excitatory ion channels that control the overall neural activity along with the brain state. In other words, mutant mice exhibit a less clear distinction in the neural activities associated with wakefull and unconscious states.

Under ethanol, MD neurons lacking Cav3.1 show no burst and a wake state-like neural activity

In order to identify the mechanism linked to Cav3.1 mutant mice ethanol-resistant phenotype, we recorded neural firing of neurons during the FWT and following a hypnotic dose of ethanol (3.0 g/kg, I.P. injection). We focused on the first loss of motion (fLOM) as it is most analogous to the classical LORR and showed the most consistency between animals 36. fLOM also illustrates best the acute effect induced by ethanol before secondary metabolization enters in play.

Under ethanol, we observed that in WT mice a majority of neurons showed burst firing mode (Fig. 4-A1; 20/33 bursting neurons). We found a significant higher burst event rate (Fig. 4-B1; p <0.0001, Ranksum test with Holm-Bonferroni correction) and in the ratio of burst-to-total spike (Fig. 4-B2; p <0.0001, Ranksum test with Holm-Bonferroni correction) comparing walk (awake active) to fLOM (unconscious, unresponding). Mutant neurons, consistently with NREM data, did not show burst firing during fLOM (0/36 bursting neurons).

Resistance to the loss of consciousness in Cav3.1 mutant is associated with maintenance of neural activity and absence of burst.

(A) Representative time plot for, from top to bottom, the normalized activity, single unit raster plot (blue and red dots for tonic and burst firing, respectively), population mean firing (spikes/sec) and burst-to-total spike ratio (%) for Cav3.1 (+/+) (A1) and Cav3.1 (-/-) mice (A2). (B) Boxplots of Cav3.1 wild and mutant burst firing rate (spikes/sec)(B1) and burst event rate (#/min)(B2) in RS neurons of the MD during FWT walk (pre I.P. injection) and during FWT first loss of motion (fLOM, post I.P. injection) for Cav3.1 (+/+) and Cav3.1 (-/-) mice; The inset numbers in B1 indicate the ratio of the number of neurons showing burst firing over total neurons. Multiple comparisons were performed using two samples ranksum test or paired sign rank test with Holm-Bonferroni correction. *** indicates a p-value < 0.001. (C) Boxplots tonic firing rate (spikes/sec) in RS neurons of the MD during FWT walk (pre I.P. injection) and during FWT first loss of motion (fLOM, post I.P. injection) for Cav3.1 (+/+) and Cav3.1 (-/-) mice. Group and brain state effect and interaction were assessed using a two-way repeated-measure ANOVA. Post-hoc multiple comparison performed using two samples rank sum test or paired sign rank test with Holm-Bonferroni correction.*** indicates a p-value < 0.001. (D) Normalized Z-score firing during fLOM with respect to wake state (home cage) firing (D1). WT and mutant distribution and cell count based on fLOM Z-score showing increase (>1.96), no change (<1.96 and >-1.96) or decrease (<-1.96) in firing in Cav3.1 (+/+) and Cav3.1 (-/-) mice (D2). (E) UMAP (uniform manifold approximation and projection) 2-dimensional representation of wakeful states (walk: + symbol; wake: x symbol; left panel), sleep states (REM: empty triangle symbol; NREM: empty round symbol; middle panel) and fLOM state (filled round symbol; right panel) of Cav3.1 (+/+) (blue symbols) and Cav3.1 (-/-) (red symbols) mice. The all-state overlay is depicted on the far right panel.

Notably, in WT, we observed that ethanol induced a significant decrease in total firing from walking to fLOM states (Fig.4-A1 and -C; p <0.0001, Rank Sum test with Holm-Bonferroni correction) and well below wakefulness level (Homecage awake state). As in sleep, we found that a majority of RS neurons showed decreased tonic firing (total number of spikes) together with an increase in burst firing, indicating a switch in firing mode under ethanol sleep. Interestingly, the mutant mice did not show a significant decrease in total firing (Fig. 4-C; p = 0.130, Ranksum test with Holm-Bonferroni correction) and showed no burst as in sleep.

We quantified the change in activity of individual neurons using Z-score normalized to the homecage wakeful state. Here, we also observed that WT RS neurons showed a significantly reduced Z-score under ethanol fLOM whereas in mutant mice cells did not (Fig. 4-D1; normalized from home cage wake state; t(62) = −5.1400, p < 0.0001, student t-test). Remarkably, we found that a majority of WT MD neurons (29/31) showed individual significantly decreased Z-scores (Fig. 4-D2; z-threshold defined from a p-value of 0.05). During fLOM, mutant RS neurons subdivided into three populations (Fig. 4-D2) with decreasing (12/33, 36.4%), maintaining (11/33, 30%) and increasing (10/33, 30.3%) activity as measured by the Z-score with respect to wakefulness (i.e. home cage wake state). These results were consistent in individual mice (Fig. S7A) and the distribution of neural population spiking (Fig. S7B), validating that significant drop in neural activity is associated with loss of movement.

Altogether, these results indicate that, in WT, ethanol induced a strong reduction in neural activity and a switch to bursting firing mode correlated with loss of consciousness. However, under ethanol, MD mutant neurons maintained their activity to a level within homecage wake state without switching to bursting. This indicates that the drop in neural activity under ethanol is modulated by Cav3.1 t-type calcium channels. In its absence, MD mutant neurons display an overall reduced activity in all brain states, however, under ethanol, remain within state wakeful levels.

Under ethanol, 20 Hz neurostimulation of MD induces mutant-like resistance to loss of consciousness in WT mice

We observed that the maintenance of neural activity in MD excitatory neurons might be at the origin of the ethanol resistance in mutant mice. We hypothesized that artificially maintaining MD neural activity within the wakeful level would sustain consciousness under ethanol. In addition, we hypothesized that the triggering of burst firing under ethanol would potentiate loss of consciousness under ethanol. To test this possibility, we used electric and optogenetic stimulations during the FWT in WT mice under an hypnotic dose of ethanol.

MD neurons in WT mice showed a spike firing range 0-50 Hz with an average neural firing around ∼20 Hz during homecage wakefulness (Fig. S7-B). Using the 20 Hz stimulation (Fig. 5-A, inter-pulse interval = 50 msec, pulse width = 6.25 msec) in MD neurons transduced with excitatory channelrhodopsin 40,41 (aav-syn-ChR2-sfGFP; Fig. 5-B and Fig. S9; see Methods), we observed an increase in ethanol resistance, which was demonstrated by a significant increase in latency to fLOM (Fig. 5-D1; Z(13) = −2.372, p = 0.013; Ranksum test). The duration of fLOM (Fig. 5-D2; Z(13) = 2.256, p = 0.020; Ranksum test) and total time spent in LOM (Fig. 5-D3; Z(13) = 2.488, p = 0.009; Ranksum test) were also significantly reduced. We verified that optogenetic stimulation of MD neurons at 20Hz (Fig. S8) induced action potentials at the same frequency with a latency response of about ∼5 ms (Fig. S8-B2; pulse width = 6.25 msec). We also observed that, although marginally higher, optogenetic stimulation did not induce any significant increase in locomotor activity (Fig. S8-C; F(1,12) = 3.6232, p = 0.0812) in the control and stimulated group which indicates that the stimulation-induced increase in ethanol resistance was not due to an increase in locomotor activity.

Optogenetic 20 Hz stimulation of MD in WT mice mimics ethanol resistance.

(A) Mice were transduced bilaterally in the MD with a AAV-syn-CHR2-sfGFP and implanted with bilateral optic fibers targeting MD with an entry angle of 30 deg relative to the sagittal plane. We used a stimulation protocol of tonic-like pulses at 20 Hz with 50 ms inter-pulse-interval (IPI) and 6.25 ms pulse duration. (B) Representative expression of ChR2-sfGFP in the MD (dashed white lines) with fiber optic ending (white arrows). (C) In vivo response of a MD RS neurons to the 20Hz tonic stimulation protocol using 6.25 ms pulse at 20Hz and laser power of 6.0 mW (C1); magnification of the peristimulus response of the neuron around the laser pulse (1 ms bin; C2). (D) Latency to first LOM (D1), duration of the first LOM (D2) and total time spent in LOM state (D3) over a recording duration of 1 hour post I.P. injection of 3.0g/Kg of ethanol are shown for the control group (No Stim) and stimulated group (Tonic-like). * is for p<0.05, ** is for p<0.01, *** is for p<0.001.

In order to validate this observation, we then bilaterally implanted mice with twisted wires for bipolar, local electric stimulation of MD (Fig. S11). As in the optogenetics experiment, we used a continuous pulse train of 20Hz electric stimulation (Fig. S10-A; upper panel; inter-pulse interval = 50 msec, pulse width = 1 msec) in addition to a burst-like stimulation (Fig. S10-A; lower pane; 4 pulses at 4 ms interval and interburst interval of 1 sec) that showed, respectively, tonic-like and burst-like entrainment in thalamic neurons 42. We observed that our 20Hz tonic-like stimulation significantly increased the latency to fLOM (Fig. S10-B1; tonic-like vs no stim.: p = 0.008; tonic-like vs burst-like: p = 0.007; Ranksum test with Holm-Bonferroni correction) and significantly decreased the total time spent in LOM (Fig. 5-B3; tonic-like vs no stim.: p = 0.008; tonic-like vs burst-like: p = 0.007, Ranksum test with Holm-Bonferroni correction). No significant changes in the duration of the first LOM was observed (duration of fLOM (Fig. 5-B2; tonic-like vs no stim.: p = 0.917; tonic-like vs burst-like: p = 0.606; Ranksum test with Holm-Bonferroni correction).

Interestingly, we observed that burst-like electrical stimulation (Fig. S10-A; lower panel; inter-spike interval = 4 msec, inter-burst interval = 1 sec, pulse width = 1 msec) did not induce any significant change in ethanol sensitivity compared to the no stimulation group (Latency to fLOM: p = 0.365; duration of fLOM: p = 0.835; total time spent in LOM: p = 1; Ranksum test with Holm-Bonferroni correction). This result suggests that burst firing alone might not have a role in ethanol resistance.

Altogether, these results suggest that the maintenance of MD firing at wakefulness level (20Hz) causally drives resistance to loss of consciousness after a hypnotic dose of ethanol. Burst-like stimulation alone did not promote or reduce loss of consciousness. This result supports the idea that neural activity maintenance in MD promotes the maintenance of consciousness even under heavy sedatives.

Discussion

In this work, we identified that the MD plays a causal role in the loss or maintenance of consciousness. The whole body Cav3.1 KO and MD-specific Cav3.1 KD resulted in the resistance to the transition to the unconscious state that was induced by hypnotic dose of ethanol in mice. In WT mice, the MD neurons demonstrated a clearly reduced firing rate in natural and ethanol-induced unconscious states compared to conscious states. This neural activity reduction was obscured in KO mice. In particular, transition to an unconscious state was accompanied with a switch of firing mode from tonic firing to burst firing in WT mice whereas this mode-shift disappeared in KO mice. Notably, the optogenetic and electric stimulation of the MD after the ethanol injection was sufficient to cause the resistance to the transition to unconscious state, supporting our conclusion that the level of neural firing in the MD is critical to maintain conscious state or resist to unconscious state, and the Cav3.1 calcium channel is causally involved in this process.

MD is a modulator of consciousness

The role of mediodorsal (MD) thalamic nucleus in perception, attention 43 and emotional control 42,44 has been the dominant focus thus far. The recent investigations on thalamic control of consciousness revealed that nuclei within dMT holds an important modulatory function in the interaction of attention and arousal 9,30. Particularly, centromedian thalamic nucleus, and not VB, showed rapid shifts in LFP preceding brain state transitions such as NREM and propofol-induced anesthesia 8. The centrolateral thalamic nucleus was implicated in the modulation of arousal and improvement of consciousness during seizure 32 and paraventricular thalamic nucleus showed critical involvement in wake/sleep cycle regulation 33. The mediodorsal thalamic nucleus, however, has rarely been included as a possible pathway in the direct modulation of consciousness 34. The MD receives projections from various parts of the basal forebrain 9 and brainstem nuclei such as the pedunculopontine nucleus that control the ascending pathway of arousal and attention 45. The MD is known to innervate limbic region, basal ganglia and medial prefrontal cortex 46 and increased activity in MD might modulate the stability of cortical UP state and synchronization 9,26. In the present study, we observed that MD (Fig. 2), but not VB KD of Cav3.1 increased ethanol resistance in mice (Fig. S3). We found that MD neurons in Cav3.1 mutant mice exhibited tonic firing within range of wakefulness (Fig. 3 and 4) and might be indicative of resistance to ethanol. In addition, we found a strong association between the normalized tonic firing in MD and the arousal level, indicating that MD tonic firing could be interpreted both as a thalamic readout and a modulator of the brain state 11 (Fig. 3). To our knowledge, this is the first report demonstrating the causal involvement of mediodorsal thalamic nucleus in the modulation of the level of consciousness in mice under anesthetic.

Cav3.1 T-type calcium channels drive thalamic firing mode and activity

The decrease of absolute firing rate observed in thalamic neurons of Cav3.1 mutant mice supports the polyvalent role of Cav3.1 in controlling both burst and tonic firing in the thalamus. Cav3.1 channels are major contributors of excitability, and in their absence or blockade, leads to a reduced neural excitability and stability and lower tonic relay of thalamocortical cells under wake-like state 47,48. The burst and tonic firing-mediated response of thalamic neurons under sensory stimulation and under the control of thalamocortical layer 6 projecting neurons was found to recruit Cav3.1 t-type calcium channels to differentiate salient novel stimuli versus complex coded information49. Therefore, the nonlinear amplification and regularization of excitatory postsynaptic potentials (EPSPs) by Cav3.1 T-type calcium channels through complexes such as with metabotropic-glutamatergic receptor 1 (mGluR1) 50 or the role of a “T window” 51 would explain how the lack of T-current in mutant mice could result in an overall reduced excitation of thalamic neurons. Cav3.1 T-type is therefore a major excitatory ion channel of the central thalamic neurons.

The lower variability in MD Firing reflects Ethanol Resistance in Cav3.1 mutant mice

Under acute hypnotic dose of ethanol, two mechanisms might favor the reduction in firing in MD: 1) an increase in synaptic and extrasynaptic GABAergic inhibition 17 and/or 2) reduced NMDA synaptic transmission 7. The presence of burst firing during fLOM, and during LOM in general, supports that MD neurons might have been subject to GABA receptor-mediated hyperpolarization, a necessary condition for the de-inactivation of Cav3.1 T-type burst. However, considering the dramatic difference in tonic firing observed during the FWT following I.P. injection of ethanol, the change in tonic firing in MD was the focus of our analysis.

We observed a reduction in neural firing under ethanol sleep conditions in WT mice (Fig. 4-C, -D1 and -D2) suggesting that low firing level should be associated with a state of low consciousness as observed during NREM sleep. Although, mutant RS neurons in MD showed an overall decrease in excitability and variability of firing in various natural conscious and unconscious states, Cav3.1 mutant mice did not exhibit a decreased arousal level (i.e. increased locomotor activity), rather they exhibited increased resistance to ethanol and increased locomotor activity, suggesting relative change from state to state in tonic firing in MD, and not the absolute value of firing, might be a better correlate of arousal in the mice. Our optogenetic and electrical stimulation showed that a sustained tonic-like stimulation in the MD at 20Hz (Fig. 5-A), a physiological relevant firing rate in wake state (Fig. 4), could increase ethanol resistance in WT mice. Reducing MD firing using phasic inhibition under ethanol, potentially leading to inhibition and rebound burst 52, could also increase the duration of the fLOM in WT mice injected with a lower dose of ethanol (Figure S12; 2.0 g/kg). We propose that the relative change in firing rate in MD RS neurons might be an important driver and indicator of the change of transition in and out of consciousness as demonstrated for other nuclei of the dMT 8,9,32,33. Therefore, the low variability in firing of MD in Cav3.1 mutant mice might be the driving force for the higher resistance to loss of motion under ethanol. In mutants, brain states might be less distinguishable leading to frequent sleep stages switch23 or resistance to unconsciousness35.

Cav3.1 T-type Calcium and Burst during Low Conscious State

Burst, as a result of Cav3.1 T-type calcium channel de-inactivation/activation, is thought to control the gating of sensory-motor stimuli 53,54 and modulate attention towards novel stimuli rather than the transmission of details 49,53,55. Previous reports highlighted the importance of burst in the stabilization of low level of consciousness 13,22,23 suggesting a direct role for burst, while no mention of the importance of tonic firing was made. We found that the propensity for burst during ethanol-induced LOM (Fig. 4-A1, -B1 and -B2; fLOM: 20/33 bursting neurons; 0.79±1.63 Burst event/min) was lower than in NREM (NREM: 34/34 bursting neurons; 5.76 ± 5.51 burst events/min) and higher than during wakefulness (Wake: 5/36 bursting neurons; 0.16±0.31 Burst event/min). In addition, burst-like electrical stimulation of MD did not significantly affect ethanol resistance (Fig. S10). Although burst-like stimulations are highly artificial and do not recruit T-current and associated mechanisms following low-threshold burst, they allow for the reproduction of the influence of TC burst firing on target centers, including thalamo-cortical and thalamo-thalamic efferents.

Interestingly, under lower doses of ethanol (I.P. injection of 2.0g/kg of ethanol) mutant and WT alike showed similar levels of resistance to ethanol. We observed a phasic inhibition of MD neurons in WT, capable of inducing partial silencing 56 and rebound bursts 57 (Fig. S12; 1sec ON-OFF using archaerhodopsin-mediated inhibition), did increase fLOM duration mostly (Z(13) = −2.214, p = 0.022, Ranksum test). This result supports that in the context of hypnotic dose of ethanol, the apparition of burst might correlate with unconscious state stability rather than induction. Burst stimulation without inhibition did not have this effect (Fig. S10 and S11). Currently, our data does not allow us to formulate any clear conclusion on the direct role of burst events during fLOM. We propose that the absence of burst and an accompanying effect of maintenance of tonic firing under ethanol in MD was responsible for the observed increase in resistance and maintenance of activity in Cav3.1 mutant mice.

A bidirectional modulation of Cav3.1 expression and alcoholism

In human mutation of Cav3.1 T-type channels exhibit mental disorders including cerebellar ataxia, absence seizure, schizophrenia and autism58. Remarkably, mutation in voltage-gated calcium channels, including Cav3.1 leads to ethanol resistance and alcohol-seeking behavior 21. Reversibly, chronic exposure to ethanol intake is known to impair sleep 59 and increase ethanol resistance. Previous studies have found an alteration in T-type calcium channel expression following chronic exposure to ethanol in non-human primates 60 suggesting that a reduced T-current and the resulting sustained thalamic tonic firing could be a possible mechanism for early stage ethanol resistance in alcoholic subjects, which increases the conversion probability from casual to compulsive consumption of ethanol. The lack of burst and sustained tonic firing might impair the stabilization of sleep, and in turn chronic sleep impairments might engage addiction related networks. Mechanisms such as adenosine receptor depreciation 61,62 or GABA-receptor potentiation 18,63,64 would enhance ethanol resistance and addiction 65,66 spiraling into further sleep fragmentation, memory consolidation deficit, impulsivity and other impairments associated with alcoholism.

eLife Assessment

This study presents a valuable finding on the relationship between neuronal dynamics in the thalamus and brain state modulation. The evidence supporting the claims of the authors is incomplete, as additional analyses and better presentation of the data are needed to support the specific role of the mediodorsal nucleus in this phenomenon. The work will be of interest to systems neuroscientists interested in brain dynamics and behavioural states.

Reviewer #1 (Public Review):

Summary:

This an interesting and valuable study that uses multiple approaches to understand the role of bursting involving voltage-gated calcium channels within the mediodorsal thalamus in the sedative-hypnotic effects of alcohol. Given its unique functional roles and connectivity pattern, the idea that the mediodorsal thalamus may have a fundamental role in regulating alcohol-induced transitions in consciousness state would be both important for researchers investigating thalamocortical dynamics and more broadly interesting for understanding brain function. In addition, the author's examination of the role of the voltage-gated calcium channel Cav3.1 provides some evidence that burst-firing mediated by this channel in the thalamus is functionally important for behavioral-state transitions. While many previous studies have suggested an analogous role for sleep-state regulation, the evidence for an analogous role of this type of bursting in sedative-induced transitions is more limited. Despite the importance of these results, however, there is some concern that the manipulations and recording approaches employed by the authors may affect other thalamic nuclei adjacent to the MD, such as the central lateral nucleus, which has also been implicated in controlling state transitions. The evidence for a specific role of the mediodorsal thalamus is therefore somewhat incomplete, and so additional validation is needed.

Strengths:

This study employs multiple, complementary research approaches including behavioral assays, sh-RNA-based localized knockdown, single-unit recordings, and patterned optogenetic interventions to examine the role of activity in the mediodorsal thalamus in the sedative-hypnotic effects of alcohol. Experiments and analyses included in the manuscript generally appear well conceived and are also generally well executed. Sample sizes are sufficiently large and statistical analysis appears generally appropriate though in some cases additional quantification would be helpful. The findings presented are novel and provide some interesting insight into the role of the thalamus as well as voltage-gated calcium channels within this region in controlling behavioral state transitions induced by alcohol. In particular, the observed effects of selective knockout along with recordings in total knockout of the voltage-gated calcium channel, Cav3.1, which has previously been implicated in bursting dynamics as well as state transitions, particularly in sleep, together suggest that the transition of thalamic neurons to a bursting pattern of firing from a more constant firing is important for transition to the sedated state produced by ethanol intoxication. While previous studies have similarly implicated Cav3.1 bursting in behavioral state transitions, the direct optogenetic interventions and single-unit recordings provide valuable new insight. These findings may also have interesting implications for the relationship between sleep process disruption associated with ethanol dependence, although the authors do not appear to examine this directly or extensively discuss these implications of their findings.

Weaknesses:

A key claim of the study is that the mediodorsal thalamus is specifically important for the sedative-hypnotic effect of ethanol and that a transition to a bursting pattern of firing in this circuit facilitates these effects due to a loss of a more constant tonic firing pattern. Despite the generally clear observed effects across the included experiments, however, the evidence presented does not fully support that the mediodorsal thalamus, in particular, is involved. This distinction is important because some previous studies have suggested that another thalamic nucleus which is very close to the mediodorsal thalamus, the central-lateral thalamus, has previously been suggested to play a role in preventing sedative-induced transitions. Despite its proximity to the mediodorsal thalamus, the central-lateral thalamus has a substantially different pattern of connectivity so distinguishing which region is impacted is important for understanding the findings in the manuscript. While sh-RNA knockdown appears to be largely centered in the mediodorsal thalamus in the example shown, (Figure 2) this is rather minimal evidence and it is also not well explained (indeed, the relevant panels do not even appear to be referenced in the text of the manuscript) and the consistency of the knockdown targeting is not quantified. Additional evidence should be provided to validate this approach. Similarly, while an example is shown for the expression of ChR2 (Fig. 5) there seems to be some spread of expression outside of the mediodorsal thalamus even in his example raising a concern about how regionally specific this effect.

The recordings targeting the mediodorsal thalamus could provide evidence of a direct association between changes in activity specifically in this part of the thalamus with the behavioral measures but there are currently some issues with making this link. One difficulty is that, although lesions are shown in Figure S5 to validate recording locations, this figure is relatively unclear and the examples appear to be taken from a different anterior/posterior location compared to the reference diagram. A larger image and improved visualization of the overall set of lesion locations that includes multiple anterior/posterior coronal sections would be helpful. Moreover, even for these example images, it is difficult to evaluate whether these are in the mediodorsal thalamus, particularly given the small size of the image shown. Ideally, an example image that is more obviously in the mediodorsal thalamus would also be included. Finally, an assessment of the relationship between the approximate locations of recorded neurons across the tetrode arrays and the behavioral measures would be very helpful in supporting the unique role of the mediodorsal thalamus. The lack of these direct links, in combination with the histological issues, reduces the insight that can be gained from this study.

In addition to the key experimental issues mentioned above, there are often problems in the text of the manuscript with reasoning or at least explanation as well as numerous minor issues with editing. The most substantial such issue is the lack of clarity in discussing the mediodorsal thalamus and other adjacent thalamic nuclei, such as the central-lateral nucleus, in the author's discussion of previous findings. Given that at last one of the manuscripts cited by the authors (Saalman, Front. Sys. Neuro. 2014) has directly claimed that central-lateral, rather than the mediodorsal, thalamus is important for arousal regulation related to a conscious state, this distinction should be addressed clearly in the discussion rather than papered over by grouping multiple thalamic nuclei as being medial. As part of this discussion, it would be important to consider additional relevant literature including Bastos et al., eLife, 2021 and Redinbaugh et al., Neuron, 2020 which are quite critical but currently do not appear to be cited. Considering additional literature relevant to the function of the mediodorsal thalamus would also be beneficial.

While the methods employed generally seem sound, the description in the methods section is lacking in detail and is often difficult to follow. Analysis methods such as the burst index appear to only be given a brief explanation in the text and appear not to be mentioned in the methods section. Similarly, the staining method used in Figure 2 does not appear to be described in the methods section. The most substantial case is for the UMAP approach used in Figure 4-E which does not appear to be described in the methods or even described in the main text. The lack of detailed descriptions makes it difficult to evaluate the applicability and quality of the experimental and analytical approaches. Citations justifying the use of methods such as the approach to separate regular spiking and narrow spiking neuron subtypes are also needed.

Beyond the problems with content and reasoning discussed above, there are also some relatively minor issues with the clarity of writing throughout the paper (for example, in the abstract the authors refer to "the ethanol resistance behavior in WT mice" but it is difficult to parse what they mean by this statement. Similarly, the next sentence "These results support that the maintenance..." while clearer, is not well phrased. Though individually minor, issues like this re-occur throughout the manuscript and sometimes make it difficult to follow so the text should be revised to correct them. There are also some problems with labels such as the labels of A1/A2 in Figure 4, which appear to be incorrect. Also, S7 has no label on the B panels. Finally, some references are not included (only a label of [ref]).

Reviewer #2 (Public Review):

In the current study, Latchoumane and collaborators focus on the Cav3.1 calcium channels in the mediodorsal thalamic nucleus as critical players in the regulation of brain-states and ethanol resistance in mice. By combining behavioural, electrophysiological, and genetic techniques, they report three main findings. First, KO Cav3.1 mice exhibit resistance to ethanol-induced sedation and sustained tonic firing in thalamocortical units. Second, knocked-down Cav3.1 mice reproduce the same behaviour when the mediodorsal, but not the ventrobasal, thalamic nucleus is targeted. Third, either optogenetic or electric stimulation of the mediodorsal thalamus reduces ethanol-induced sedation in control animals.

Overall, the study is well designed and performed, correctly controlled for confounds, and properly analysed. Nonetheless, it is important to address some aspects of the report. The results support the conclusions of the study. These results are likely to be relevant in the field of systems neuroscience, as they increase the molecular evidence showing how the thalamus regulates brain states.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This is an interesting and valuable study that uses multiple approaches to understand the role of bursting involving voltage-gated calcium channels within the mediodorsal thalamus in the sedative-hypnotic effects of alcohol. Given its unique functional roles and connectivity pattern, the idea that the mediodorsal thalamus may have a fundamental role in regulating alcohol-induced transitions in consciousness state would be both important for researchers investigating thalamocortical dynamics and more broadly interesting for understanding brain function. In addition, the author's examination of the role of the voltage-gated calcium channel Cav3.1 provides some evidence that burst-firing mediated by this channel in the thalamus is functionally important for behavioral-state transitions. While many previous studies have suggested an analogous role for sleep-state regulation, the evidence for an analogous role of this type of bursting in sedative-induced transitions is more limited. Despite the importance of these results, however, there is some concern that the manipulations and recording approaches employed by the authors may affect other thalamic nuclei adjacent to the MD, such as the central lateral nucleus, which has also been implicated in controlling state transitions. The evidence for a specific role of the mediodorsal thalamus is therefore somewhat incomplete, and so additional validation is needed.

Strengths:

This study employs multiple, complementary research approaches including behavioral assays, sh-RNAbased localized knockdown, single-unit recordings, and patterned optogenetic interventions to examine the role of activity in the mediodorsal thalamus in the sedative-hypnotic effects of alcohol. Experiments and analyses included in the manuscript generally appear well conceived and are also generally well executed. Sample sizes are sufficiently large and statistical analysis appears generally appropriate though in some cases additional quantification would be helpful. The findings presented are novel and provide some interesting insight into the role of the thalamus as well as voltage-gated calcium channels within this region in controlling behavioral state transitions induced by alcohol. In particular, the observed effects of selective knockout along with recordings in total knockout of the voltage-gated calcium channel, Cav3.1, which has previously been implicated in bursting dynamics as well as state transitions, particularly in sleep, together suggest that the transition of thalamic neurons to a bursting pattern of firing from a more constant firing is important for transition to the sedated state produced by ethanol intoxication. While previous studies have similarly implicated Cav3.1 bursting in behavioral state transitions, the direct optogenetic interventions and single-unit recordings provide valuable new insight. These findings may also have interesting implications for the relationship between sleep process disruption associated with ethanol dependence, although the authors do not appear to examine this directly or extensively discuss these implications of their findings.

Weaknesses:

A key claim of the study is that the mediodorsal thalamus is specifically important for the sedative-hypnotic effect of ethanol and that a transition to a bursting pattern of firing in this circuit facilitates these effects due to a loss of a more constant tonic firing pattern. Despite the generally clear observed effects across the included experiments, however, the evidence presented does not fully support that the mediodorsal thalamus, in particular, is involved. This distinction is important because some previous studies have suggested that another thalamic nucleus which is very close to the mediodorsal thalamus, the central-lateral thalamus, has previously been suggested to play a role in preventing sedative-induced transitions. Despite its proximity to the mediodorsal thalamus, the central-lateral thalamus has a substantially different pattern of connectivity so distinguishing which region is impacted is important for understanding the findings in the manuscript. While sh- RNA knockdown appears to be largely centered in the mediodorsal thalamus in the example shown, (Figure 2) this is rather minimal evidence and it is also not well explained (indeed, the relevant panels do not even appear to be referenced in the text of the manuscript) and the consistency of the knockdown targeting is not quantified. Additional evidence should be provided to validate this approach. Similarly, while an example is shown for the expression of ChR2 (Fig. 5) there seems to be some spread of expression outside of the mediodorsal thalamus even in his example raising a concern about how regionally specific this effect.

The recordings targeting the mediodorsal thalamus could provide evidence of a direct association between changes in activity specifically in this part of the thalamus with the behavioral measures but there are currently some issues with making this link. One difficulty is that, although lesions are shown in Figure S5 to validate recording locations, this figure is relatively unclear and the examples appear to be taken from a different anterior/posterior location compared to the reference diagram. A larger image and improved visualization of the overall set of lesion locations that includes multiple anterior/posterior coronal sections would be helpful. Moreover, even for these example images, it is difficult to evaluate whether these are in the mediodorsal thalamus, particularly given the small size of the image shown. Ideally, an example image that is more obviously in the mediodorsal thalamus would also be included. Finally, an assessment of the relationship between the approximate locations of recorded neurons across the tetrode arrays and the behavioral measures would be very helpful in supporting the unique role of the mediodorsal thalamus. The lack of these direct links, in combination with the histological issues, reduces the insight that can be gained from this study.

In addition to the key experimental issues mentioned above, there are often problems in the text of the manuscript with reasoning or at least explanation as well as numerous minor issues with editing. The most substantial such issue is the lack of clarity in discussing the mediodorsal thalamus and other adjacent thalamic nuclei, such as the central-lateral nucleus, in the author's discussion of previous findings. Given that at last one of the manuscripts cited by the authors (Saalman, Front. Sys. Neuro. 2014) has directly claimed that central-lateral, rather than the mediodorsal, thalamus is important for arousal regulation related to a conscious state, this distinction should be addressed clearly in the discussion rather than papered over by grouping multiple thalamic nuclei as being medial. As part of this discussion, it would be important to consider additional relevant literature including Bastos et al., eLife, 2021 and Redinbaugh et al., Neuron, 2020 which are quite critical but currently do not appear to be cited. Considering additional literature relevant to the function of the mediodorsal thalamus would also be beneficial. While the methods employed generally seem sound, the description in the methods section is lacking in detail and is often difficult to follow. Analysis methods such as the burst index appear to only be given a brief explanation in the text and appear not to be mentioned in the methods section. Similarly, the staining method used in Figure 2 does not appear to be described in the methods section. The most substantial case is for the UMAP approach used in Figure 4-E which does not appear to be described in the methods or even described in the main text. The lack of detailed descriptions makes it difficult to evaluate the applicability and quality of the experimental and analytical approaches. Citations justifying the use of methods such as the approach to separate regular spiking and narrow spiking neuron subtypes are also needed.

Beyond the problems with content and reasoning discussed above, there are also some relatively minor issues with the clarity of writing throughout the paper (for example, in the abstract the authors refer to "the ethanol resistance behavior in WT mice" but it is difficult to parse what they mean by this statement. Similarly, the next sentence "These results support that the maintenance..." while clearer, is not well phrased. Though individually minor, issues like this re-occur throughout the manuscript and sometimes make it difficult to follow so the text should be revised to correct them. There are also some problems with labels such as the labels of A1/A2 in Figure 4, which appear to be incorrect. Also, S7 has no label] on the B panels. Finally, some references are not included (only a label of [ref]).

Reviewer #2 (Public Review):

In the current study, Latchoumane and collaborators focus on the Cav3.1 calcium channels in the mediodorsal thalamic nucleus as critical players in the regulation of brain-states and ethanol resistance in mice. By combining behavioural, electrophysiological, and genetic techniques, they report three main findings. First, KO Cav3.1 mice exhibit resistance to ethanol-induced sedation and sustained tonic firing in thalamocortical units. Second, knocked-down Cav3.1 mice reproduce the same behaviour when the mediodorsal, but not the ventrobasal, thalamic nucleus is targeted. Third, either optogenetic or electric stimulation of the mediodorsal thalamus reduces ethanol-induced sedation in control animals.

Overall, the study is well designed and performed, correctly controlled for confounds, and properly analysed. Nonetheless, it is important to address some aspects of the report. The results support the conclusions of the study. These results are likely to be relevant in the field of systems neuroscience, as they increase the molecular evidence showing how the thalamus regulates brain states.

Reviewer #1 (Recommendations For The Authors):

Aside from the additional quantification and clarification of the analysis discussed in the weakness section, in general, the experiments included in the manuscript seem reasonable. However, I would suggest one additional experiment as well as one control, both of which are relatively straightforward optogenetic experiments, that I feel would be helpful to further improve the study. First, as the authors note, the optogenetic interventions used do not directly address the relevance of the changes in bursting patterns observed in the knockout (KO), which are by far the most robust effect, with the changes in alcohol sensitivity. One approach that could help address this would be to use patterned suppression via inhibitory opsins (e.g. halorhodopsin) to "rescue" the periods of inhibition associated with bursting in the KO. Localizing this inhibition to the mediodorsal thalamus would also lend further credence to their claim that this nuclei is the relevant circuit for their observed effects. For the control, tonic activation of the ventrobasal nucleus, as the authors did for the mediodorsal nucleus, would be beneficial to rule out the possibility that the observed effect would occur with any thalamic nucleus. In addition to these experiments, I did not note the strategy for sharing data obtained through this study so this should be added.

R1 – 1: A key claim of the study is that the mediodorsal thalamus is specifically important for the sedative-hypnotic effect of ethanol and that a transition to a bursting pattern of firing in this circuit facilitates these effects due to a loss of a more constant tonic firing pattern. Despite the generally clear observed effects across the included experiments, however, the evidence presented does not fully support that the mediodorsal thalamus, in particular, is involved. This distinction is important because some previous studies have suggested that another thalamic nucleus which is very close to the mediodorsal thalamus, the central-lateral thalamus, has previously been suggested to play a role in preventing sedative-induced transitions. Despite its proximity to the mediodorsal thalamus, the central-lateral thalamus has a substantially different pattern of connectivity so distinguishing which region is impacted is important for understanding the findings in the manuscript.

R1-A1: The reviewer is right that CL has been pointed as another candidate structure with causal influence on arousal and consciousness. We have focused our efforts in including only recording single units that were from tetrode located in the MD specifically using the lesion code we explain in the method section and in response to R1 question#3. We also produced a quantification of Cav3.1 knock-down that clearly demonstrates that the KD experiment was itself specific to MD, bilaterally, and that CL to CM were minimally impacted by the knock-down process (Fig. 2C and D). Moreover, the optogenetic (fiber incidence was 30 degrees guaranteeing a central coverage rather than lateral; Fiber optic NA = 0.22) and electric stimulation (bipolar twisted electrodes, 50uA) experiments were also very selective and specific to the MD (Fig.S5). It remains clear that MD might not be the sole structure involved in the brain state control towards sedation and “anesthetic states”, and CL might be a significant contributor as well, however, we show that CL manipulations were rather irrelevant in our experiments (Fig. 2, S5, S9 and S11).

R1-2: While sh-RNA knockdown appears to be largely centered in the mediodorsal thalamus in the example shown, (Figure 2) this is rather minimal evidence and it is also not well explained (indeed, the relevant panels do not even appear to be referenced in the text of the manuscript) and the consistency of the knockdown targeting is not quantified. Additional evidence should be provided to validate this approach.

R1-A2: In order to address this important question, we have created an additional panel quantification to fig2D. We have then quantified the intensity per area of Cav3.1 expression in sub zones of 4 regions of interest: MD (left, right; 2 subzones each), Centro Medial (CM; 1 subzones in total), Centrolateral/Paraventricular nucleus (CL/PCN; left, right; 2 subzones each) and the submedial nucleus (SMT; left, right; used as a control for the intensity normalization; 1 subzones in total). This panel clearly illustrates that MD was knocked-down bilaterally (p<0.001). Moreover, CM (p<0.05) and CL (p<0.01) were also partially and unilaterally knocked down, as well. This analysis confirms that our KD had a high specificity to MD.

We added the relevant figure caption and text:

[Result section, Cav3.1 silencing in the MD, but not VB, increased ethanol resistance in mice, paragraph 3]

“We then characterized the change in Cav3.1 expression following the shControl and shCav3.1 knockdown injections in three test regions MD (left and right), CM (centromedial nucleus) and CL (centrolateral nuclei, left and right side) and a negative control region SMT (submedial thalamic nuclei, left and right side). The average intensity was obtained from two coronal brain slices for each mice used in the experiment (see Methods sections, Cav3.1 Intensity quantification). Our results show that the targeting of the knockdown was very specific to the bilateral MD (p<0.001; Fig. 2D). We noted that the CM (p<0.05) and a marginal unilateral knock-down of the CL were also observed (p<0.01). Notably, we tested the correlation between the level of knock-down in MD and the total time in LOM and observed a significant association (Fig. 2D inset; R = 0.599, p = 0.018). This result highlights that the Cav3.1 knock-down was specific to MD and with an intensity associated with ethanol-induced loss of motion.”

R1-3: One difficulty is that, although lesions are shown in Figure S5 to validate recording locations, this figure is relatively unclear and the examples appear to be taken from a different anterior/posterior location compared to the reference diagram. A larger image and improved visualization of the overall set of lesion locations that includes multiple anterior/posterior coronal sections would be helpful. Moreover, even for these example images, it is difficult to evaluate whether these are in the mediodorsal thalamus, particularly given the small size of the image shown. Ideally, an example image that is more obviously in the mediodorsal thalamus would also be included. Finally, an assessment of the relationship between the approximate locations of recorded neurons across the tetrode arrays and the behavioral measures would be very helpful in supporting the unique role of the mediodorsal thalamus.

R1-A3: Related to fig.S5, we re-distributed the position of the recordings from the tetrode electrode burned positions over 3 representative coronal planes that best represent the implant positions. We also provided additional snapshots of tetrode location. To identify the positions of four tetrodes in each animal, we encoded the positions with different electrical lesion strategies as follows: 1 lesion(tetrode 1), 2 lesions while we redrew the tetrode with 100 um interval (tetrode 2), 3 lesions with 200um interval (tetrode 3), 4 lesions with 50um intervals (tetrode4). Tetrodes that were found outside of the MD delimited region were discarded post analysis. A straight relationship between the closeness of the electrode is unfortunately not possible for tetrode recording, a straight silicone probe which maintains the spatial spacing in recording would have been a better approach in that case, but unfortunately, it was not performed in our study.

R1-4: In addition to the key experimental issues mentioned above, there are often problems in the text of the manuscript with reasoning or at least explanation as well as numerous minor issues with editing. The most substantial such issue is the lack of clarity in discussing the mediodorsal thalamus and other adjacent thalamic nuclei, such as the central-lateral nucleus, in the author's discussion of previous findings. Given that at last one of the manuscripts cited by the authors (Saalman, Front. Sys. Neuro. 2014) has directly claimed that central-lateral, rather than the mediodorsal, thalamus is important for arousal regulation related to a conscious state, this distinction should be addressed clearly in the discussion rather than papered over by grouping multiple thalamic nuclei as being medial. As part of this discussion, it would be important to consider additional relevant literature including Bastos et al., eLife, 2021 and Redinbaugh et al., Neuron, 2020 which are quite critical but currently do not appear to be cited. Considering additional literature relevant to the function of the mediodorsal thalamus would also be beneficial.

R1-A4: We thank the reviewer for his comments and suggestions. We agree that the added references mentioned by the reviewers are highly relevant and should be integrated in the manuscript. We have integrated the above-mentioned references and further developed on the discussion on the role of MD relative to other thalamic nuclei (ILN and CL in particular). We believe that this better-referenced and clarified text does improve the manuscript greatly.

[introduction section, paragraph 3]

“The centrolateral (CL) thalamic nucleus has been implicated in the modulation of arousal, behavior arrest 31, and improvement of level of consciousness during seizures 32. Notably, the direct electrical stimulation of the intralaminar nuclei (ILN) and, in particular CL, promoted hallmarks of arousal and awakening in primate under propofol and ketamine propofol anesthesia.”

[Discussion section, paragraph 1]

“In this work, we identified that the neural activity in MD plays a causal role in the maintenance of consciousness. Whole body Cav3.1 KO and MD-specific Cav3.1 KD mice showed resistance to loss of consciousness induced by hypnotic dose of ethanol. In WT mice, MD neurons demonstrated a reduced firing rate in natural (sleep) and ethanol-induced unconscious states compared to awake states. This neural activity reduction was impaired in KO mice. In particular, transition to an unconscious state was accompanied with a switch of firing mode from tonic firing to burst firing in WT mice whereas this modeshift disappeared in KO mice. Finally, optogenetic or electric stimulations of the MD after ethanol injection were sufficient to induce a resistance to loss of motion, supporting that the level of neural firing in the MD is critical to maintain conscious state and delay unconscious state. We showed that the expression of Cav3.1 t-type calcium channels in MD is a cellular modulator associated with this effect.”

[Discussion section, MD is a modulator of consciousness, paragraph 2 and 3]

“The MD is known to innervate limbic region, basal ganglia and medial prefrontal cortex 50 and increased activity in MD might modulate the stability of cortical UP states (e.g. awaken, aroused and attentive states) and synchronization 9,26. Thus, MD might be a major hub involved in cortical state control and brain state stabilization.

Supporting the brain state stabilization theory and the ethanol resistance of Cav3.1 mutants, Choi et al.34 demonstrated that the loss of Cav3.1 T-type calcium channel reduced the bilateral coherence between PFC and MD under ketamine anesthesia and ethanol hypnosis, especially in the delta frequency bands. More importantly, under propofol anesthesia, Bastos et al.35 showed that intralaminar nucleus and MD stimulation lead to increased wake-up subscore and arousal, together with an increased in cortico-cortico and thalamo-cortical slow (delta) frequency power.

In the present study, we observed that MD KD (Fig. 2A), but not VB KD (Fig. S3) of Cav3.1 increased and is associated (Fig. 2D) with ethanol resistance in mice. We found that MD neurons in Cav3.1 mutant mice exhibited tonic firing within range of wakefulness (Fig. 3 and 4), indicative of resistance to ethanol and wake-like brain state. In addition, we found a strong association between the normalized tonic firing in MD and the arousal through brain states (i.e. walk to wake to sleep states), supporting that MD tonic firing could be interpreted both as a thalamic readout and a modulator of the brain state 11 (Fig. 3). Finally, direct optogenetic and electric MD stimulation increased resistance to loss of consciousness in WT mice (Fig.5 and Fig. S10). To our knowledge, this is the first report demonstrating the causal involvement of mediodorsal thalamic nucleus in the modulation of wakefulness and the resistance to ethanol-induced loss of consciousness in mice.”

R1-5: While the methods employed generally seem sound, the description in the methods section is lacking in detail and is often difficult to follow. Analysis methods such as the burst index appear to only be given a brief explanation in the text and appear not to be mentioned in the methods section.

R1-A5: We have added a clear definition in the supplementary method following the original work used:

[Supplementary Method section, Single Unit recording, sorting and analysis, last paragraph]

“The bursting index was derived as described in (Royer et al. 2012). Namely, the burst index was estimated from the spike auto-correlogram (1-ms bin size) by subtracting the mean value between 40 and 50 ms (baseline) from the peak measured between 0 and 10 ms. Positive burst amplitudes were normalized to the peak and negative amplitudes were normalized to the baseline to obtain indexes ranging from −1 to 1.” We also edited its mention in the text for clarity:

[Result section, Lack of Ca3.1 in MD neurons removes thalamic burst in NREM sleep, paragraph 2]

“[…] and a clear reduction in total bursting represented as bursting index (Fig. 3-B; ratio of spikes count <10 ms and >50 ms based on auto-cross-correlogram).”

R1-6: Similarly, the staining method used in Figure 2 does not appear to be described in the methods section.

R1-A6: The staining method can be found in the supplementary method of the paper. [supplementary method, Immunohistochemistry]

R1-7: The most substantial case is for the UMAP approach used in Figure 4-E which does not appear to be described in the methods or even described in the main text.

R1-A7: Regarding the method, the UMAP approach is described in the supplementary method document [Uniform Manifold Approximation and Projection (UMAP)]. We believe that only a succinct description was needed here considering the extent of the analysis. Regarding the inserts in the main text, we agree that the main text was lacking the description of these results and we have amended the main text to reflect a clear description of this result and what it entails. The following paragraph was added:

[Result section, Under ethanol, MD neurons lacking Cav3.1 show no burst and a wake state-like neural activity, second to last paragraph]

“Finally, we asked whether the firing modes and properties (tonic firing rate, burst firing rate; see supplementary methods) of single MD neurons would form distinct qualitative representation of “brain stages” using a lowered dimensional UMAP representation (Uniform Manifold Approximation and Projection42 ). We observed that for awake and active (i.e. walk), the brain state representation formed two adjacent clusters that confounded both wild and mutant neurons (Fig. 4E, left panel). The REM and NREM states, the wild type neurons formed 2 additional interconnected clusters, whereas the mutant neurons tend to overlap with the clusters attributed to the “awake” brain state (Fig. 4E, second to left panel). Ethanol induced fLOM, similarly to REM and NREM clusters, was distinct from awake clusters in wild type mice and overlapped with the NREM clusters (Fig. 4E, third to left panel). Here also, mutant MD neurons showed overlap with the awake clusters rather than the “low consciousness” brain states. These results indicate that the firing mode and properties could define a brain state representation that shows distinctions in levels of consciousness. Moreover, the mutant showed a representation of “low consciousness” states overlapping with wild type “awake” states consistent with the hypothesis of resistance to loss of consciousness.”

R1-8: Citations justifying the use of methods such as the approach to separate regular spiking and narrow spiking neuron subtypes are also needed.

R1-A8: We have added two references related to the observation of the two subpopulations of spiking neurons [Schiff and Reyes, 2012; Destexhe, 2008].

R1-9: Beyond the problems with content and reasoning discussed above, there are also some relatively minor issues with the clarity of writing throughout the paper (for example, in the abstract the authors refer to "the ethanol resistance behavior in WT mice" but it is difficult to parse what they mean by this statement.

R1-A9: We addressed this issue by editing and revising the manuscript for clarity and flow.

R1-10: Similarly, the next sentence "These results support the maintenance..." while clearer, is not well phrased. Though individually minor, issues like this re-occur throughout the manuscript and sometimes make it difficult to follow so the text should be revised to correct them.

R1-A10: We thank the reviewer for highlighting this point. We have edited the overall text to improve clarity and flow.

[abstract]

These results suggest that maintaining MD neural firing at a wakeful level is sufficient to induce resistance to ethanol-induced hypnosis in WT mice.

R1-11: There are also some problems with labels such as the labels of A1/A2 in Figure 4, which appear to be incorrect.

R1-A11: We noted this issue and have rectified the figure for clarity.

R1-12: Also, S7 has no label on the B panels.

R1-A12: We thank the reviewer for pointing out this lack. We have added the y-label on the panel for clarity.

R1-13: Finally, some references are not included (only a label of [ref]).

R1-A13: We have completed the missing reference and thank the reviewer for pointing that out.

Additional comments

R1-14: Aside from the additional quantification and clarification of the analysis discussed in the weakness section, in general, the experiments included in the manuscript seem reasonable. However, I would suggest one additional experiment as well as one control, both of which are relatively straightforward optogenetic experiments, that I feel would be helpful to further improve the study. First, as the authors note, the optogenetic interventions used do not directly address the relevance of the changes in bursting patterns observed in the knockout (KO), which are by far the most robust effect, with the changes in alcohol sensitivity. One approach that could help address this would be to use patterned suppression via inhibitory opsins (e.g. halorhodopsin) to "rescue" the periods of inhibition associated with bursting in the KO.

R1-A14: Here the reviewer proposes an interesting experiment which we have attempted to perform, however, poses several technical challenges. First, the KO do not have burst firing as they are depleted from Cav3.1 low-threshold calcium channel. Therefore, under ethanol, even if there might exist a rhythmic inhibition that activates Cav3.1 channels and causes a rebound burst, the KO are unable to have it. Therefore, an optogenetic inhibition would only accentuate the total inhibition and could potentially induce an overall decrease in MD firing, resulting in an increase in LOM features. Alternatively, we showed that in a WT with low ethanol dose (where LOM induction is harder), the increased rhythmic inhibition does indeed increase significantly LOM duration and marginally decreases latency to LOM (Fig. S12), indicating that increased inhibition could indeed explain the hypothesis: “ the stronger the decrease in MD firing, the faster and longer the LOM.” The only caveat of using WT here is that optogenetic inhibition might also include rebound burst post-inhibition. Injecting bursts only did not alter the response to ethanol (Fig. S10). These results point to the role of loss of firing in MD as a main factor for LOM, and potentially the contribution of burst necessitating a concurrent inhibition/loss of firing.

We agree that inhibition in KO would further validate this hypothesis, controlling for the role of burst. We regret that we are not in the capacity to perform additional experiments involving the KO mice.

R1-15: For the control, tonic activation of the ventrobasal nucleus, as the authors did for the mediodorsal nucleus, would be beneficial to rule out the possibility that the observed effect would occur with any thalamic nucleus.

R1-A15: We agree with the reviewer that we could have added an additional region control to the gain/loss of function experiments. We would even go further as to suggest that a better control nucleus would be a high order nucleus such as PO or an unrelated sensory relay nucleus such as LGN. VB being a motor relay nucleus, could also mediate movement initiation, which could be hard to interpret. Since the complete control study for all thalamic nuclei Cav3.1 KD is outside the scope of this study, we opted not to redo these experiments and keep the focus of the manuscript on the manipulation of MD activity rather than the various available thalamic nuclei. We also do not claim that MD is the sole center able to initiate a switch in the loss of consciousness, and a more in-depth study on that matter would be clearly needed.

R1-16: In addition to these experiments, I did not note the strategy for sharing data obtained through this study so this should be added.

R1-A16: We have uploaded data and code for most figures at the following repository and provided a clearer statement regarding data sharing. We thank the reviewer for pointing out this missing element.

The link for the repository is the following:

It contains:

- Excel spreadsheet file of all behavior values, including the newly quantified Cv3.1 expression in MD/CL/SMT

- Excel spreadsheet follow-up of all MD cells (single unit; tetrode) analyzed

- Folders for all groups studied with representative figures showing EEG power over time and normalized activity (WT vs KO for 2, 3 and 4 g/kg; MDshKD vs shCTR, VBshKD vs shCTR; CHR2 NOSTIM vs STIM; ESTIM Groups and ARCH NOSTIM vs STIM)

- A1G LORRvsLOM and OPEN FIELD Matlab data

- Matlab and ImageJ Codes: single unit analysis, characterization, brain state characterization, sleep stages, LOM, open field analysis and statistical analysis.

We have added the data sharing subsection in the acknowledgements:

“Part of the analyzed data and codes are available on the open access platform, mendeley:

Latchoumane, Charles-francois (2024), “Mediodorsal thalamic nucleus mediates resistance to ethanol through Cav3.1 T-type Ca2+ regulation of neural activity”, Mendeley Data, V1, doi: 10.17632/7fr427426m.1

Additional data (large size recording and images) can be provided upon reasonable requests.”

Reviewer #2 (Recommendations For The Authors):

R2-1. Consciousness is a contentious subject. Even in humans, there is still intense research on the topic, not to mention animals, about which we still know very little. Moreover, consciousness is not quantified in this study, as there is no standard metric to do so. Accordingly, talking about 'modulation', 'transition', ́level ', or 'reduction' of consciousness can be misleading. Hence, it is probably safer to strictly refer to brain-states and/or stages of the sleep-wake cycle in this study and reframe it entirely around these concepts.

R2-A1. The reviewer points to an important point and we appreciate this highlight. Agreeing that the definition of consciousness is rather loose and arguably difficult to pinpoint. Here, we settle on a definition that relies on the loss of motion and loss of righting reflex. This definition is widely accepted as the “verified” state in which the absence of responsiveness (to continuous stimuli, inducing reflex or discomfort) is observed and uninterrupted by jerks and spurious movements. Additional metrics needed would be the recording of EMG to quantify atonia and EEG to the settling of a dominantly slow-wave frequency (~4 Hz; ethanol-induced sedation at theta rhythm), as shown in Fig S1A. The driver of this 4Hz frequency and its correlation has been investigated previously (e.g. Choi et al, PNAS, 2012), leading to the accepted link between LOM/LORR and loss of consciousness. Our data present the advantage of showing single neuron recordings and that LOM is a state where the lowest firing activity is present (Fig S7AB) and comparable to deep sleep state activity (Fig3D). The first LOM is the most important as it highlights the deepest loss of consciousness before the ethanol starts to be metabolized and cleared, which would be consistent between animals.

As a result, we have edited the manuscript to clarify all mentions related to brain states and states of unconsciousness.

R2-2. It is not clear why the authors focus on the mediodorsal nucleus. This should be better explained in the introduction and developed in the discussion.

R2-A2. This comment converges with the Reviewer 1 comments and we are addressing this lack in the discussion as suggested. We have addressed it with this previous comment and believe it is now clearer.

R2-3. The discussion mentions that 'increased activity in MD might modulate the stability of cortical UP state and synchronization' (pg 21). This point should be either further developed and put into context, or removed. In its current state, it does not seem to contribute much to the discussion of results.

R2-A3. We understand that the working “UP state” might not be clear enough. We have modified this sentences as follows to clarify that UP state could be either a state of where the animal is awake, aroused or attentive:

[Discussion section, MD is a modulator of consciousness, first paragraph]

“The MD is known to innervate limbic region, basal ganglia and medial prefrontal cortex 50 and increased activity in MD might modulate the stability of cortical UP states (e.g. awaken, aroused and attentive states) and synchronization 9,26. Thus, MD might be a major hub involved in cortical state control and brain state stabilization.“

R2-4. The discussion states that 'mutant mice did not exhibit a decreased arousal level (i.e. increased locomotor activity)' (pg 23). This is confusing as decreased arousal should be reflected in decreased locomotor activity.

R2-A4. We understand that the formulation of this sentence may be confusing and we have edited this portion of the text to improve quality in the revised version of the manuscript. To clarify, mutant mice do not exhibit reduced or increased arousal (not quantified, just observational), they do have a phenotypic hyperlocomotion. This comes in contrast with a lower basal firing rate in the MD, which in our interpretation, is not synonymous with lower arousal. We believe that the relative change in MD determines the change in arousal, and that the absolute firing is not indicative of arousal in itself, only in comparison.

[Discussion section, The lower variability in MD Firing reflects Ethanol Resistance in Cav3.1 mutant mice, paragraph 2]

“Mutant RS neurons in MD showed an overall lower excitability and variability of firing in various natural conscious and unconscious states compared to wild type mice. Remarkably, Cav3.1 mutant mice exhibited a clear increased locomotor activity and an increased resistance to ethanol. The general lower firing rate and the high “arousal” observed in mutant mice suggests that the relative change from state to state in tonic firing in MD, and not the absolute value of firing, might be a better correlate of change in brain state in the mice.”

R2-5. The methods (pg 27) state that two genetic backgrounds (129/svjae and C57BL/6J ) were used in the study. Authors should show whether there were significant differences between those backgrounds in the key parameters assessed in the study (particularly resistance to ethanol sedation).

R2-A5. As mentioned in the method section, we only used the F1-background mice, which are the firstgeneration offspring produced by crossing 129/svjae and C57BL/6J strains. To produce F1 KO mice, we kept the heterozygote mice in two strains. We unfortunately did not study the particular difference of the respective KO of these two backgrounds; however, the pure C57BL/6J KO has been used in other studies by our group (Kim et al 2001; Na et al, 2008; Park et al., 2010). The F1 background allows us to work with mice that are less aggressive and can be handled with less inherent stress.

R2-6. It would be convenient to produce a supplementary figure associated with Figure 1C to show the same data with averages per mouse. That is, 9 points for control and 9 points for KO mice. This also applies to all cases where data is not presented per mouse but pooled between animals.

R2-A6. We have added a panel C in Figure S1, to show the scatter values for all the mice corresponding to the figure 1C. We have also generalized this presentation for all behavior graphics showing all the animals in the scatter plot next to the boxplot. We believe that this presentation increases further the transparency of the manuscript. We have then added the scatter plot for all mice in figure Fig1, Fig2, Fig5, Fig.S2, Fig.S3, Fig.S10 and Fig.S12.

R2-7. It would be informative to make a supplementary figure associated with Figure 1D to compare baseline raw activity levels (i.e., baseline walking recording) between control and KO mice. That is, do KO and control mice cover comparable distances and at similar speeds during baseline conditions? Figure 1D and Figure 4A suggest that the variability of locomotor activity is larger in KO mice. Hence, this parameter should be quantified and reported.

R2-A7. We thank the reviewer for this comment. We strived to answer to this question in the manuscript in two ways:

- We first measure the overall hyperlocomotion of the mice using the open field total distance parkoured in our mice cohorts (FigS4C). We did observe that the KO mutant showed hyperlocomotion, but not MD or VB knock-down mice. Which indicates that the hyperlocomotion component is not specific to the two thalamic nuclei studied.

- Using the forced walking task, we impose on the animal to keep a steady pace of roughly 6cm/s. This assay allows to normalize the general walking behavior to a relatively fixed pace making it comparable for all animals.

The reviewer suggested reporting the mean and variance in walking of WT and KO during baseline (prior to the ethanol I.P. injection). We believe that the two points mentioned above are sufficient to describe in a more quantitative way the WT vs KO locomotion differences. Moreover, by construction the normalized locomotion on the forced walking task will return similar means for the baseline, the standard deviation would, however, potentially show differences but would remain inconclusive.

R2-8. The legend in Figure 1 states that 'the loss of consciousness is evaluated using normalized moving index using either video analysis (differential pixel motion), on- head accelerometer-based motion, or neck electromyograms'. Authors should clarify whether these methods are equivalent and support it with data.

R2-A8. We understand the reviewer point and we have made a few modifications to the method description aligning better with what was done. For most mice, video analysis was used to obtain the moving index. When video recording was not available (2 mice), we had an accelerometer attached to the animal’s head stage which helped us derive a moving index that was similar to the video moving index. The neck electromyogram was rather used for animals implanted with the tetrodes to identify sleep stages based on local field potential frequency and muscle tone. We have then clarified the method for this matter and Figure 1 to avoid this confusion. Since no concurrent recording of both video and accelerometer was performed, we do not have the data to compute the correlation between the two measures, however, no noticeable deviation from loss of motion was observed between the two methods. We realize that this may be a weak argument, however, our observations showed that video and accelerometers returned very similar timings for loss of motion (only a few comparative instances insufficient to present a statistical comparison).

R2-9. How were spike bursts defined? The authors should try different criteria and verify the consistency of results.

R2-A9 For in vivo single unit recording, we opted for a definition that is validated from our works and others as a silencing of at least 100 ms followed by a minimum of 3 spikes with:

- First spike pairs interspike interval less than 4 ms

- Remaining spike pairs interspike interval less than 20 ms

We have performed this analysis using a minimum of 2 spikes, and varied silencing periods between 50 and 100ms, without observing significant deviation of the results. As shown in Figure S6B, with this approach we observed that the burst distribution had a majority with <10 spikes per burst. Figure S6C indicated that with a clear distribution of ISI for first spike within 2-4ms as observed in previous works (Desai and Varela, 2021; Alitto et al, 2019), importantly, not clearly capped at 4 ms, showing that the range for the first ISI might indeed be lower than 4ms for thalamic burst. Within burst spike waveforms can become very variable and the choice of 3 over 2 spikes minimum per burst stems from the aim to reduce false positive detection of ultra-short bursts, which in single unit recording remains controversial (Gray et al. 1995).

Minor:

R2-10: Figure 4A2 'Cav3.1(+/+)' should presumably be Cav3.1(-/-).

R2-A10: this is correct and we have corrected the figure label [This sentence is ambiguous. What is ‘this’ that is correct?]

R2-11: Figure S2C legend states 'Post-hoc group comparison was performed using.' The sentence seems to be incomplete.

R2-A11: We have completed the sentence for clarity.

R2-12: In the methods (pg 29) virus concentration is reported as '107 TU/ul', which probably refers to 10e7.

R2-A12: We have corrected it by superscripting the power 7.

R2-13: Verify Fig 1C1 and correct Y-axis overlap between title and units.

R2-A13: We edited the figure for clarity, thank you.

R2-14: On page 24 there is a '[ref]' that probably stands for (a missing) reference.

R2-A14: the missing reference has been added.

Version 2

Abstract

Abstract

Thalamocortical activity is known to orchestrate sensory gating and consciousness switching. The precise thalamic regions involved, or the firing patterns related to the unconsciousness remain unclear. Interestingly, the highly-expressed thalamic T-type calcium currents have been considered as a candidate for the ionic mechanism for the generation of thalamo-cortically-driven change in conscious state. Here, we tested the hypothesis that Cav3.1 T-type channels in the mediodorsal thalamic nucleus (MD) might control neuronal firing during unconsciousness using Cav3.1 T-type channel knock-out (KO) and knock-down (KD) mice under natural sleep and ethanol-induced unconsciousness. During natural sleep, the MD neurons in KO mice showed general characteristics of sustained firing across sleep stages. We found that KO and MD-specific KD mice showed enhanced resistance to ethanol. During ethanol-induced unconscious state, wild-type (WT) MD neurons showed a significant reduction in neuronal firing from baseline with increased burst firing, whereas Cav3.1 KO neurons showed well sustained neural firing, within the level of wakefulness, and no burst firing. Further, 20 Hz optogenetic and electrical activation of MD neurons mimicked the ethanol resistance behavior in WT mice. These results suggest that maintaining MD neural firing at a wakeful level is sufficient to induce resistance to ethanol-induced hypnosis in WT mice. This work has important implications for the design of treatments for consciousness disorders using thalamic stimulation of deeper nuclei including the targeting of the mediodorsal thalamic nucleus.

Introduction

Drug-induced unconsciousness can be achieved using numerous types of anesthetics with varying modes of action 1,2. Ethanol, one of the most frequently abused drugs in human society, can induce sleep-like loss of consciousness at high doses 3. While possible neuropharmacological and neural correlates of ethanol sedation have been proposed using in vitro and in vivo methods 4–7, recent studies have highlighted the slowing of thalamocortical-driven rhythms as a potent marker of unconsciousness 8,9. However, the region and the mechanism linked to the thalamic modulation during ethanol induced unconsciousness remains poorly understood.

Physiological correlates of thalamocortical rhythmic activities and consciousness state of the brain have been studied extensively10–13. T-type calcium channels are known generators of thalamocortical rhythms, through the modulation of cell excitability and rebound burst firing14,15. During sleep, the transition from wakefulness to unconsciousness is associated with membrane hyperpolarization of thalamic neurons 12,16. Similarly, It has been shown that in ethanol sedation, as in natural sleep or absence seizure, the loss of consciousness is characterized by a switch from tonic to burst firing in thalamic neurons, which involves GABAergic inhibition-driven de-inactivation of Cav3.1 T-type channels resulting in slow oscillatory response of the thalamocortical network 7,17–19. Acute intoxication at high doses of ethanol 20,21 induces both slow oscillations in the delta-theta frequency range and a loss of righting reflex (LORR), a classical proxy to assess the loss of consciousness. It has been shown that mice lacking global or thalamic Cav3.1 showed altered slow oscillations and sleep architecture 22,23; delayed sleep induction under several anesthetics (i.e. isoflurane, halothane, sevoflurane and pentobarbital) 24; and increased resistance to drug-induced absence seizure 13. Notably, the absence or blockade of Cav3.1 resulted in an increased preference for ethanol consumption and novelty-seeking behavior 20,21. In the current studies we investigate the role in brain state modulation of Cav3.1-mediated T-currents during ethanol-induced sleep.

The thalamus is one of the major regions expressing Cav3.1 T-type calcium channels 25 and holds a central role in information-transmission and integration 26. In vitro and in vivo studies using genetically modified mice have revealed that Cav3.1 T-type channels play a key role in the genesis of thalamocortical rhythms, such as 3Hz spike-and-wave discharges (SWD), a signature of absence seizures 13,27 and delta waves 16,28,29. Previous investigations on thalamic control of consciousness revealed that nuclei within the dorsal medial thalamus (dMT) hold an important modulatory function in the interaction of attention and arousal 9,30. Particularly, centromedian (CM) thalamic nucleus, and not ventrobasal nucleus (VB), showed rapid shifts in local field potential (LFP) preceding brain state transitions such as NREM and propofol-induced anesthesia 8. The paraventricular thalamic nucleus (PVN) showed critical involvement in wake/sleep cycle regulation 33. The centrolateral (CL) thalamic nucleus has been implicated in the modulation of arousal, behavior arrest 31, and improvement of level of consciousness during seizures 32. Notably, the direct electrical stimulation of the intralaminar nuclei (ILN) and, in particular CL, promoted hallmarks of arousal and awakening in primate under propofol and ketamine propofol anesthesia. The MD, a sub nucleus of dMT, on the other hand, has only recently been implicated in disorders of consciousness 33 and ketamine/ethanol-induced loss of consciousness34 through the alteration of thalamo-cortical functional connectivity. In anesthetized primate, the stimulation of ILN and MD increased arousal and wakefulness score35. However, several key questions still remain to be answered: 1) Is there a specific role for MD Cav3.1 T-type calcium channels in the control of ethanol-induced loss of consciousness? 2) Whether Cav3.1 T-type calcium channel driven neuronal firing pattern has any role in the control of consciousness?

In this study, we identified that KO and MD specific silencing of Cav3.1 T-type calcium channels result in increased ethanol resistance in mice. Using single unit recordings, we compared MD activity of WT and KO mice while the mice transitioned from conscious to unconscious state and found that the KO mice showed more sustained MD activity whereas the WT mice showed clearly reduced MD activity. Furthermore and consistently with their resistant phenotype, KO mice showed sustained MD firing, well within the wakefulness level, under ethanol consumption. Finally, we demonstrate that both the optogenetic and electrical stimulations in MD, mimicking the sustained firing pattern of knock-out mice, were sufficient to induce the increased ethanol resistance in WT mice. These results reveal a causal control of brain state by MD during ethanol-induced unconsciousness and with an underlying neural mechanism governed by Cav3.1 T-type calcium channels.

Results

To understand the role of T-type Ca2+ channels in modulating the consciousness level, we compared the ethanol resistance between WT and Cav3.1−/− KO littermates. We used the forced walking task (FWT; Fig. 1-A), an analog to the LORR assay, which enables a continuous and high-temporal resolution assessment of the loss of movement 36 (LOM). Moreover, the FWT objectively measures the latency to and duration of the first LOM, but also the total time spent in LOM using automatized analysis of video confirmed by electromyograms (EMGs) or accelerometer recordings (Fig. 1-B and S1; see Methods). The continuously running treadmill (6 cm/s) ensures a normalized behavior within and between animals before injection (i.e., baseline forced walking) and allows for reduced intervention from experimenters for the monitoring of both electrophysiological (Fig. S1, upper panel) and analyzed behavioral (Fig. S1, lower panel).

Mice lacking Cav3.1 showed increased ethanol resistance on the forced walking task.

(A) The schematic of the FWT setup. Mice are habituated and trained on a constantly moving treadmill (6 cm/sec). Following a baseline walking recording (∼10 min), the mouse is carefully picked up and injected with ethanol (i.p.). Once placed back on the treadmill, the loss of consciousness is evaluated using normalized moving index using either video analysis (differential pixel motion), on-head accelerometer- based motion, or neck electromyograms over a period of 60 min.

(B) Representative EEG (parietal) and 3-axis accelerometer (Acc) traces for walking and LOM in a Cav3.1 (+/+) mouse.

(C) Quantification for the latency to first LOM (fLOM; i.e. delay between I.P. Injection and loss of motion; C1), the duration of the fLOM (C2) and the total time spent in LOM state (C3) over a recording duration of 60 min following 3.0 g/Kg I.P injection of ethanol in Cav3.1 (+/+) and Cav3.1 (-/-) mice; data is represented as boxplot wish individual mice as scatter plot. * is for p<0.05, ** is for p<0.01 and *** is for p<0.001.

(D) Representative normalized motor activity over time for Cav3.1(+/+) (D1) and Cav3.1 (-/-) (D2) mice post I.P. injection of 3.0 g/Kg. Blue and red boxes above the graph indicate the state interpretation for walking and LOM, respectively.

The lacking Cav3.1 increases ethanol resistance in mice

Mice lacking Cav3.1 exhibited delayed anesthetic induction 13,24 and impairment in maintenance of low conscious level 22,23 as well as increased ethanol preference 21,34. We tested the sensitivity of Cav3.1 null mutant mice for various acute hypnotic doses of ethanol.

We confirmed that the lack of Cav3.1 resulted in a more delayed and fragmented LOM (Fig. 1-C1 and C2), and a reduction in the total time spent in LOM compared to Cav3.1 WT mice (Fig. 1-C3). We observed that Cav3.1 null mutant mice showed increased latency to and decreased duration of the first episode of loss of motion (fLOM) for ethanol injection doses of 2.0, 3.0 and 4.0 g/Kg (Fig. S2). The total time spent in LOM during one hour of recording was also significantly reduced (Fig. S2-A3 and -B3) compared with WT mice. Two-way analysis of variance (ANOVA) showed a significant effect for the main factor genotype and dose for the latency to and duration of fLOM, and total time in LOM (Table S1), indicating a dose dependency in both wild and mutant mice. In particular, an I.P. Injection of 3.0 g/Kg induced a significant difference between wild and mutant mice in latency to (t(16) =-4.1965, p = 0.002; Student t-test) and duration of fLOM (t(16) =2.3908, p = 0.0294; Student t-test), and total time spent in LOM (t(16) =3.9065, p = 0.0012; Student t-test).

These results indicate that ethanol induces a dose-dependent sedative effect on mice and Cav3.1 mutant mice had an increased resistance to ethanol sedation compared to WT mice.

Cav3.1 silencing in the MD, but not VB, increased ethanol resistance in mice

A great majority of Cav3.1 expression is found in the thalamic region 25 and was shown to specifically correlate with the modulation of thalamocortical-related rhythms and stability of sleep level 22,23. To identify a possible role of the thalamic region in ethanol resistance, we knocked down the expression of Cav3.1 in the MD and a ventral basal nucleus (VB) region, two regions possibly involved in a thalamic control of consciousness 33,34,37, using a lentivirus (LV)-mediated short hairpin (shRNA) delivery.

We found that compared to shControl injected mice, shCav3.1 knock-down of MD resulted in an increased latency to (Fig. 2-A1; t(27) = -3.0045, p = 0.0057; Student t-test) and duration of (Fig. 2-A2; t(27) =2.1448, p = 0.0411; Student t-test) fLOM, and total time spent in LOM (Fig. 2-A3; t(27) =2.6641, p = 0.0128, two-tailed test) for 3.0g/Kg I.P. Injection of ethanol. However, we found that compared to shControl-injected mice, Cav3.1 KD of VB did not change the latency to (t(8) =-1.0093 p = 0.3423, two-tailed test), duration of (t(8) =-0.0983, p = 0.9241, two-tailed test) fLOM and total time spent in LOM (t(8) =- 0.6317, p = 0.5452, two-tailed test) for the same 3.0g/Kg I.P. Injection of ethanol (Fig. S3). Representative traces of mice activity showed that mice with Cav3.1 KD in MD (Fig. S4- A) had a more delayed and fragmented early period of LOM compared to MD LV- shControl and VB injected mice, as in mutant mice.

Cav3.1 knock-down in MD increased ethanol resistance in mice.

(A) Quantification for the Latency to first LOM (fLOM; A1), the duration of the first LOM (A2) and the total time spent in LOM state (A3) over a recording duration of 60 min post I.P. injection of ethanol (3.0g/Kg) in lentivirus-shControl and shCav3.1 knock-down mice for MD ; data is represented as a boxplot with individual mice shown as a scatter plot. * and ** indicates p<0.05 and p<0.01, respectively.

(B) Representative brain coronal section stained using Cav3.1 antibody (B1) showing the endogenous thalamic expression of Cav3.1 in the MD of lentivirus(LV)-shControl injected mice; Cav3.1 and GFP merging (B2) and higher magnification of the white dashed square in B2 (B3). Scale bars in (B2) and (B3) indicate 500 μm and 100 μm, respectively.

(C) Representative brain coronal section showing the reduced Cav3.1 expression in the MD of LV-shCav3.1 injected mice (C1); Cav3.1 and GFP merging (C2) and higher magnification of the white dashed square in C2 (C3). Scale bars in (C2) and (C3) indicate 500 μm and 100 μm, respectively.

(D) Normalized Cav3.1 intensity estimated for the nuclei MD, CM (centromedial), CL/PCN (Centrolateral/paracentral) and SMT (submedial thalamic nucleus). The quantification was performed as intensity per area for 2 replicates per side per mouse. *, ** and *** indicates p<0.05, p<0.01 and p<0.001, respectively (2 sample t-test). The data is shown as a scatter plot for all values and superposed with the mean and standard deviation error bars. Inset: We noted a positive correlation between the total LOM duration and the Cav3.1 intensity in MD (R = 0.599, p = 0.018).

We then characterized the change in Cav3.1 expression following the shControl and shCav3.1 knock-down injections in three test regions MD (left and right), CM (centromedial nucleus) and CL (centrolateral nuclei, left and right side) and a negative control region SMT (submedial thalamic nuclei, left and right side). The average intensity was obtained from two coronal brain slices for each mice used in the experiment (see Methods sections, Cav3.1 Intensity quantification). Our results show that the targeting of the knock-down was very specific to the bilateral MD (p<0.001; Fig. 2D). We noted that the CM (p<0.05) and a marginal unilateral knock-down of the CL were also observed (p<0.01). Notably, we tested the correlation between the level of knock-down in MD and the total time in LOM and observed a significant association (Fig. 2D inset; R = 0.599, p = 0.018). This result highlights that the Cav3.1 knock-down was specific to MD and with an intensity associated with ethanol-induced loss of motion.

During the open field test, Cav3.1 null mutant mice showed significantly increased locomotor activity compared to WT mice as shown by total distance moved (Fig. S4-C; analysis of variance: GROUP F(3) = 8.45, p = 0.0004; Cav3.1+/+ vs Cav3.1-/-: p = 0.0001). The mice with MD-specific Cav3.1 KD, however, did not show any significant difference in total distance moved compared to shControl-injected control mice (MD LV-shControl vs MD LV-shCav3.1: p = 0.868; Holm-Sidak correction), indicating that the ethanol resistance in MD Cav3.1 KD mice was not attributed by hyperlocomotion observed in Cav3.1 KO mice.

Lack of Ca3.1 in MD neurons removes thalamic burst in NREM sleep

Thalamic neurons are known to follow a state-dependent activity 38 ; however, the nature of this state-dependent activity has not been studied for the MD. In order to understand the relationship between MD neuron firing and level of consciousness, we investigated the association between the neural activity in MD and brain states at different levels of consciousness (Fig. 3 and S5).

Lack of Cav3.1 removed burst firing, and reduces neural activity and its variability in MD across natural conscious and unconscious states.

(A) Mice were implanted unilaterally with 4 tetrode wires to record single unit activity in the MD while in home cage (wake, sleep: NREM, REM) and forced walking task under ethanol (walk).

(B - Left panel) Scatter distribution of spike width vs bursting index of MD regular spiking (RS, round shape) and narrow spiking (NS, triangle shape) neurons. Cav3.1 (+/+) and Cav3.1 (-/-) neurons are marked as filled and empty shapes, respectively . The histogram of the pooled Cav3.1 (+/+) and Cav3.1 (-/-) distribution is projected on each axis. (B - Right panel) Representative auto-cross correlograms of a RS neuron showing the presence and absence of fast spiking interval (burst firing) in Cav3.1 wild-type and mutant mice, respectively.

(C) Boxplots of Cav3.1 wild and mutant burst firing rate (spikes/sec; burst spikes-only averaged over a state duration) (C1) and burst event rate (#/min; number of burst event averaged over a state duration; see burst definition in methods section) (C2) in RS neurons of the MD during NREM sleep, a stage known for the presence of bursting firing mode in thalamic neurons, for Cav3.1 (+/+) and Cav3.1 (-/-) mice; The inset numbers in C1 indicate the number of neurons showing burst firing (more than 1 event in 10 min) over the total number of single neurons identified.

(D) Boxplots of Cav3.1 wild and mutant tonic firing rate (spikes/sec) in RS neurons of the MD during walking (FWT), wake (home cage), REM and NREM sleep (home cage) for Cav3.1 (+/+) and Cav3.1 (-/-) mice. Group and brain state effect and interaction were assessed using a two-way repeated-measure ANOVA. For post-hoc, two samples Ranksum test comparison *, ** and *** indicate p-value lower than 0.05, 0.01 and 0.001, respectively. For two sample Levene’s test for homoscedasticity #, ## and ### indicate p-value lower than 0.05, 0.01 and 0.001, respectively. Pearson Ranksum correlations between brain state and total firing for Cav3.1(+/+) and Cav3.1(-/-) is indicated above boxplots.

We observed two major populations of neuronal spike waveform present in MD single unit recordings of Cav3.1 (+/+) mice, also described in previous works 39,40 : 1) a majority of regular spiking (RS) cells characterized by wide spike waveform (36/39 neurons, 92.3%), i.e. spike-to-valley width >250 μsec (Fig. 3B) and high bursting propensity; 2) a minority of narrow spiking (NS; also known as fast spiking) cells showing short spike-to- valley width, i.e. <250 μsec, lower bursting characteristics (Fig. 3 and S6-A3 and -A4) and fast-paced tonic firing (10-50Hz; data not shown). The RS and NS neurons were found in MD of WT and mutant mice; however, MD RS mutant neurons showed an absence of short inter-spike intervals (ISI, i.e. indicative of the absence of burst) in auto cross- correlogram (Fig. S6) and a clear reduction in total bursting represented as bursting index (Fig. 3-B; ratio of spikes count <10 ms and >50 ms based on auto-cross-correlogram). Since RS cells have the profile of the major population of MD, i.e. excitatory neurons, we focused on the analysis of RS neurons mainly in the remainder of this study.

During the deep sleep state non-rapid eye movement (NREM), thalamic neural firing is known to switch from tonic to burst firing 41. We found that a lack of Cav3.1 T-type calcium channels resulted in a near absence of burst (see Methods for definition) in mutant mice (Fig. 3-C1 and C2; 4/44 bursting neurons; Z(77) = 7.20, p < 0.0001, Ranksum test) compared to WT mice (34/34 bursting neurons; 5.76 ± 5.51 burst events/min).

Lack of Ca3.1 reduces neuronal activity across all brain states in MD

In addition, we observed that the mutant mice showed a significant lower total firing rate (main factor group: F(1,186) = 16.5, p = 0.0001; interaction group x brain state: F(3,186) = 4.72, p = 0.0034) and a reduced variability (p <0.0001 for all brain states; Levene’s test) in most brain states compared to the wild type mice, indicating that the lack of Cav3.1 t- type channels results in an overall reduction in neural activity in RS neurons.

RS neurons of MD in Cav3.1 wild type and mutant mice showed a significant change in overall firing across walking, waking (home cage), NREM and REM sleep states as shown by a repeated measure ANOVA (Fig. 3-D; main factor brain state: F(3,186) = 104.96, p <0.0001). In addition, Cav3.1 wild type (R = -0.534, p = 1.6e-10, Spearman ranked correlation) and mutant (R = -0.689, p = 6.8e-20, Spearman ranked correlation) showed a significant negative correlation between neural firing and brain state. Assuming an ordering from higher to lower state of consciousness, these results indicate that MD firing is associated with level of consciousness independently from the Cav3.1 T-type channels in wild type and mutant mice. Importantly, this result indicates that Cav3.1 t-type calcium channels are critical excitatory ion channels that control the overall neural activity along with the brain state. In other words, mutant mice exhibit a less clear distinction in the neural activities associated with wakeful and unconscious states.

Under ethanol, MD neurons lacking Cav3.1 show no burst and a wake state-like neural activity

In order to identify the mechanism linked to Cav3.1 mutant mice ethanol-resistant phenotype, we recorded neural firing of neurons during the FWT and following a hypnotic dose of ethanol (3.0 g/kg, I.P. injection). We focused on the first loss of motion (fLOM) as it is most analogous to the classical LORR and showed the most consistency between animals 36. fLOM also illustrates best the acute effect induced by ethanol before secondary metabolization enters in play.

Under ethanol, we observed that in WT mice a majority of neurons showed burst firing mode (Fig. 4-A1; 20/33 bursting neurons). We found a significant higher burst event rate (Fig. 4-B1; p <0.0001, Ranksum test with Holm-Bonferroni correction) and in the ratio of burst-to-total spike (Fig. 4-B2; p <0.0001, Ranksum test with Holm-Bonferroni correction) comparing walk (awake active) to fLOM (unconscious, unresponsive). Mutant neurons, consistently with NREM data, did not show burst firing during fLOM (0/36 bursting neurons).

Resistance to the loss of consciousness in Cav3.1 mutant is associated with maintenance of neural activity and absence of burst.

(A) Representative time plot for, from top to bottom, the normalized activity, single unit raster plot (blue and red dots for tonic and burst firing, respectively), population mean firing (spikes/sec) and burst-to-total spike ratio (%) for Cav3.1 (+/+) (A1) and Cav3.1 (-/-) mice (A2).

(B) Boxplots of Cav3.1 wild and mutant burst firing rate (spikes/sec)(B1) and burst event rate (#/min)(B2) in RS neurons of the MD during FWT walk (pre I.P. injection) and during FWT first loss of motion (fLOM, post I.P. injection) for Cav3.1 (+/+) and Cav3.1 (-/-) mice; The inset numbers in B1 indicate the ratio of the number of neurons showing burst firing over total neurons. Multiple comparisons were performed using two samples Ranksum test or paired sign rank test with Holm-Bonferroni correction. *** indicates a p-value < 0.001.

(C) Boxplots tonic firing rate (spikes/sec) in RS neurons of the MD during FWT walk (pre I.P. injection) and during FWT first loss of motion (fLOM, post I.P. injection) for Cav3.1 (+/+) and Cav3.1 (-/-) mice. Group and brain state effect and interaction were assessed using a two-way repeated-measure ANOVA. Post-hoc multiple comparison performed using two samples rank sum test or paired sign rank test with Holm- Bonferroni correction.*** indicates a p-value < 0.001.

(D) Normalized Z-score firing during fLOM with respect to wake state (home cage) firing (D1). WT and mutant distribution and cell count based on fLOM Z-score showing increase (>1.96), no change (<1.96 and >-1.96) or decrease (<-1.96) in firing in Cav3.1 (+/+) and Cav3.1 (-/-) mice (D2).

UMAP (uniform manifold approximation and projection) 2-dimensional representation of wakeful states (walk: + symbol; wake: x symbol; left panel), sleep states (REM: empty triangle symbol; NREM: empty round symbol; middle panel) and fLOM state (filled round symbol; right panel) of Cav3.1 (+/+) (blue symbols) and Cav3.1 (-/-) (red symbols) mice. The all-state overlay is depicted on the far right panel.

Notably, in WT, we observed that ethanol induced a significant decrease in total firing from walking to fLOM states (Fig.4-A1 and -C; p <0.0001, Rank Sum test with Holm- Bonferroni correction) and well below wakefulness level (Home cage awake state). As in sleep, we found that a majority of RS neurons showed decreased tonic firing (total number of spikes) together with an increase in burst firing, indicating a switch in firing mode under ethanol sleep. Interestingly, the mutant mice did not show a significant decrease in total firing (Fig. 4-C; p = 0.130, Ranksum test with Holm-Bonferroni correction) and showed no burst as in sleep.

We quantified the change in activity of individual neurons using Z-score normalized to the home cage wakeful state. Here, we also observed that WT RS neurons showed a significantly reduced Z-score under ethanol fLOM whereas in mutant mice cells did not (Fig. 4-D1; normalized from home cage wake state; t(62) = -5.1400 , p < 0.0001, student t-test). Remarkably, we found that a majority of WT MD neurons (29/31) showed individual significantly decreased Z-scores (Fig. 4-D2; z-threshold defined from a p-value of 0.05). During fLOM, mutant RS neurons subdivided into three populations (Fig. 4-D2) with decreasing (12/33, 36.4%), maintaining (11/33, 30%) and increasing (10/33, 30.3%) activity as measured by the Z-score with respect to wakefulness (i.e. home cage wake state). These results were consistent in individual mice (Fig. S7A) and the distribution of neural population spiking (Fig. S7B), validating that significant drop in neural activity is associated with loss of movement.

Finally, we asked whether the firing modes and properties (tonic firing rate, burst firing rate; see supplementary methods) of single MD neurons would form distinct qualitative representation of “brain stages” using a lowered dimensional UMAP representation (Uniform Manifold Approximation and Projection42 ). We observed that for awake and active (i.e. walk), the brain state representation formed two adjacent clusters that confounded both wild and mutant neurons (Fig. 4E, left panel). The REM and NREM states, the wild type neurons formed 2 additional interconnected clusters, whereas the mutant neurons tend to overlap with the clusters attributed to the “awake” brain state (Fig. 4E, second to left panel). Ethanol induced fLOM, similarly to REM and NREM clusters, was distinct from awake clusters in wild type mice and overlapped with the NREM clusters (Fig. 4E, third to left panel). Here also, mutant MD neurons showed overlap with the awake clusters rather than the “low consciousness” brain states. These results indicate that the firing mode and properties could define a brain state representation that shows distinctions in levels of consciousness. Moreover, the mutant showed a representation of “low consciousness” states overlapping with wild type “awake” states consistent with the hypothesis of resistance to loss of consciousness.

Altogether, these results indicate that, in WT, ethanol induced a strong reduction in neural activity and a switch to bursting firing mode correlated with loss of consciousness. However, under ethanol, MD mutant neurons maintained their activity to a level within home cage wake state without switching to bursting. This indicates that the drop in neural activity under ethanol is modulated by Cav3.1 t-type calcium channels. In its absence, MD mutant neurons display an overall reduced activity in all brain states, however, under ethanol, remain within state wakeful levels.

Under ethanol, 20 Hz neurostimulation of MD induces mutant-like resistance to loss of consciousness in WT mice

We observed that the maintenance of neural activity in MD excitatory neurons might be at the origin of the ethanol resistance in mutant mice. We hypothesized that artificially maintaining MD neural activity within the wakeful level would sustain consciousness under ethanol. In addition, we hypothesized that the triggering of burst firing under ethanol would potentiate loss of consciousness under ethanol. To test this possibility, we used electric and optogenetic stimulations during the FWT in WT mice under a hypnotic dose of ethanol.

MD neurons in WT mice showed a spike firing range 0-50 Hz with an average neural firing around ∼20 Hz during home cage wakefulness (Fig. S7-B). Using the 20 Hz stimulation (Fig. 5-A, inter-pulse interval = 50 msec, pulse width = 6.25 msec) in MD neurons transduced with excitatory channelrhodopsin 43,44 (aav-syn-ChR2-sfGFP; Fig. 5-B and Fig. S9; see Methods), we observed an increase in ethanol resistance, which was demonstrated by a significant increase in latency to fLOM (Fig. 5-D1; Z(13) = -2.372, p = 0.013; Ranksum test). The duration of fLOM (Fig. 5-D2; Z(13) = 2.256, p = 0.020; Ranksum test) and total time spent in LOM (Fig. 5-D3; Z(13) = 2.488, p = 0.009; Ranksum test) were also significantly reduced. We verified that optogenetic stimulation of MD neurons at 20Hz (Fig. S8) induced action potentials at the same frequency with a latency response of about ∼5 ms (Fig. S8-B2; pulse width = 6.25 msec). We also observed that, although marginally higher, optogenetic stimulation did not induce any significant increase in locomotor activity (Fig. S8-C; F(1,12) = 3.6232, p = 0.0812) in the control and stimulated group which indicates that the stimulation-induced increase in ethanol resistance was not due to an increase in locomotor activity.

Optogenetic 20 Hz stimulation of MD in WT mice mimics ethanol resistance.

(A) Mice were transduced bilaterally in the MD with an AAV-syn-CHR2-sfGFP and implanted with bilateral optic fibers targeting MD with an entry angle of 30 deg relative to the sagittal plane. We used a stimulation protocol of tonic-like pulses at 20 Hz with 50 ms inter-pulse-interval (IPI) and 6.25 ms pulse duration.

(B) Representative expression of ChR2-sfGFP in the MD (dashed white lines) with fiber optic ending (white arrows).

(C) In vivo response of a MD RS neurons to the 20Hz tonic stimulation protocol using 6.25 ms pulse at 20Hz and laser power of 6.0 mW (C1); magnification of the peristimulus response of the neuron around the laser pulse (1 ms bin; C2). Latency to first LOM (D1), duration of the first LOM (D2) and total time spent in LOM state (D3) over a recording duration of 1 hour post I.P. injection of 3.0g/Kg of ethanol are shown for the control group (No Stim) and stimulated group (Tonic-like). * is for p<0.05, ** is for p<0.01, *** is for p<0.001.

In order to validate this observation, we then bilaterally implanted mice with twisted wires for bipolar, local electric stimulation of MD (Fig. S11). As in the optogenetics experiment, we used a continuous pulse train of 20Hz electric stimulation (Fig. S10-A; upper panel; inter-pulse interval = 50 msec, pulse width = 1 msec) in addition to a burst-like stimulation (Fig. S10-A; lower pane; 4 pulses at 4 ms interval and interburst interval of 1 sec) that showed, respectively, tonic-like and burst-like entrainment in thalamic neurons 45. We observed that our 20Hz tonic-like stimulation significantly increased the latency to fLOM (Fig. S10-B1; tonic-like vs no stim.: p = 0.008; tonic-like vs burst-like: p = 0.007; Ranksum test with Holm-Bonferroni correction) and significantly decreased the total time spent in LOM (Fig. 5-B3; tonic-like vs no stim.: p = 0.008; tonic-like vs burst-like: p = 0.007, Ranksum test with Holm-Bonferroni correction). No significant changes in the duration of the first LOM were observed (duration of fLOM (Fig. 5-B2; tonic-like vs no stim.: p = 0.917; tonic-like vs burst-like: p = 0.606; Ranksum test with Holm-Bonferroni correction).

Interestingly, we observed that burst-like electrical stimulation (Fig. S10-A; lower panel; inter-spike interval = 4 msec, inter-burst interval = 1 sec, pulse width = 1 msec) did not induce any significant change in ethanol sensitivity compared to the no stimulation group (Latency to fLOM: p = 0.365; duration of fLOM: p = 0.835; total time spent in LOM: p = 1; Ranksum test with Holm-Bonferroni correction). This result suggests that burst firing alone might not have a role in ethanol resistance.

Altogether, these results suggest that the maintenance of MD firing at wakefulness level (20Hz) causally drives resistance to loss of consciousness after a hypnotic dose of ethanol. Burst-like stimulation alone did not promote or reduce loss of consciousness. This result supports the idea that neural activity maintenance in MD promotes the maintenance of consciousness even under heavy sedatives.

Discussion

In this work, we identified that the neural activity in MD plays a causal role in the maintenance of consciousness. Whole body Cav3.1 KO and MD-specific Cav3.1 KD mice showed resistance to loss of consciousness induced by hypnotic dose of ethanol. In WT mice, MD neurons demonstrated a reduced firing rate in natural (sleep) and ethanol- induced unconscious states compared to awake states. This neural activity reduction was impaired in KO mice. In particular, transition to an unconscious state was accompanied with a switch of firing mode from tonic firing to burst firing in WT mice whereas this mode- shift disappeared in KO mice. Finally, optogenetic or electric stimulations of the MD after ethanol injection were sufficient to induce a resistance to loss of motion, supporting that the level of neural firing in the MD is critical to maintain conscious state and delay unconscious state. We showed that the expression of Cav3.1 t-type calcium channels in MD is a cellular modulator associated with this effect.

MD is a modulator of consciousness

The role of mediodorsal (MD) thalamic nucleus in perception, attention 46 and emotional control 45,47 has been the dominant focus thus far. The recent investigations on thalamic control of consciousness revealed that nuclei within dMT holds an important modulatory function in the interaction of attention and arousal 9,30. Particularly, centromedian thalamic nucleus, and not VB, showed rapid shifts in LFP preceding brain state transitions such as NREM and propofol-induced anesthesia 8. The centrolateral thalamic nucleus was implicated in the modulation of arousal and improvement of consciousness during seizure32 and paraventricular thalamic nucleus showed critical involvement in wake/sleep cycle regulation 48 . The mediodorsal thalamic nucleus, however, has rarely been included as a possible pathway in the direct modulation of consciousness 33. The MD receives projections from various parts of the basal forebrain 9 and brainstem nuclei such as the pedunculopontine nucleus that control the ascending pathway of arousal and attention 49. The MD is known to innervate limbic region, basal ganglia and medial prefrontal cortex 50 and increased activity in MD might modulate the stability of cortical UP states (e.g. awaken, aroused and attentive states) and synchronization 9,26. Thus, MD might be a major hub involved in cortical state control and brain state stabilization.

Supporting the brain state stabilization theory and the ethanol resistance of Cav3.1 mutants, Choi et al.34 demonstrated that the loss of Cav3.1 T-type calcium channel reduced the bilateral coherence between PFC and MD under ketamine anesthesia and ethanol hypnosis, especially in the delta frequency bands. More importantly, under propofol anesthesia, Bastos et al.35 showed that intralaminar nucleus and MD stimulation lead to increased wake-up subscore and arousal, together with an increased in cortico- cortico and thalamo-cortical slow (delta) frequency power.

In the present study, we observed that MD KD (Fig. 2A), but not VB KD (Fig. S3) of Cav3.1 increased and is associated (Fig. 2D) with ethanol resistance in mice. We found that MD neurons in Cav3.1 mutant mice exhibited tonic firing within range of wakefulness (Fig. 3 and 4), indicative of resistance to ethanol and wake-like brain state. In addition, we found a strong association between the normalized tonic firing in MD and the arousal through brain states (i.e. walk to wake to sleep states), supporting that MD tonic firing could be interpreted both as a thalamic readout and a modulator of the brain state 11 (Fig. 3). Finally, direct optogenetic and electric MD stimulation increased resistance to loss of consciousness in WT mice (Fig.5 and Fig. S10). To our knowledge, this is the first report demonstrating the causal involvement of mediodorsal thalamic nucleus in the modulation of wakefulness and the resistance to ethanol-induced loss of consciousness in mice.

Cav3.1 T-type calcium channels drive thalamic firing mode and activity

The decrease of absolute firing rate observed in thalamic neurons of Cav3.1 mutant mice supports the polyvalent role of Cav3.1 in controlling both burst and tonic firing in the thalamus. Cav3.1 channels are major contributors of excitability, and in their absence or blockade, leads to a reduced neural excitability and stability and lower tonic relay of thalamocortical cells under wake-like state 51,52. The burst and tonic firing-mediated response of thalamic neurons under sensory stimulation and under the control of thalamocortical layer 6 projecting neurons was found to recruit Cav3.1 t-type calcium channels to differentiate salient novel stimuli versus complex coded information 53. Therefore, the nonlinear amplification and regularization of excitatory postsynaptic potentials (EPSPs) by Cav3.1 T-type calcium channels through complexes such as with metabotropic-glutamatergic receptor 1 (mGluR1) 54 or the role of a “T window” 55 would explain how the lack of T-current in mutant mice could result in an overall reduced excitation of thalamic neurons. Cav3.1 T-type is therefore a major excitatory ion channel of the central thalamic neurons.

The lower variability in MD Firing reflects Ethanol Resistance in Cav3.1 mutant mice

Under acute hypnotic dose of ethanol, two mechanisms might favor the reduction in firing in MD: 1) an increase in synaptic and extra synaptic GABAergic inhibition 17 and/or 2) reduced NMDA synaptic transmission 7. The presence of burst firing during fLOM, and during LOM in general, supports that MD neurons might have been subject to GABA receptor-mediated hyperpolarization, a necessary condition for the de-inactivation of Cav3.1 T-type burst. However, considering the dramatic difference in tonic firing observed during the FWT following I.P. injection of ethanol, the change in tonic firing in MD was the focus of our analysis.

We observed a reduction in neural firing under ethanol sleep conditions in WT mice (Fig. 4-C, -D1 and -D2) suggesting that low firing level should be associated with a state of low consciousness as observed during NREM sleep. Mutant RS neurons in MD showed an overall lower excitability and variability of firing in various natural conscious and unconscious states compared to wild type mice. Remarkably, Cav3.1 mutant mice exhibited a clear increased locomotor activity and an increased resistance to ethanol. The general lower firing rate and the high “arousal” observed in mutant mice suggests that the relative change from state to state in tonic firing in MD, and not the absolute value of firing, might be a better correlate of change in brain state in the mice. Our optogenetic and electrical stimulation showed that a sustained tonic-like stimulation in the MD at 20Hz (Fig. 5-A), a physiological relevant firing rate in wake state (Fig. 4), could increase ethanol resistance in WT mice. Reducing MD firing using phasic inhibition under ethanol, potentially leading to inhibition and rebound burst 56, could also increase the duration of the fLOM in WT mice injected with a lower dose of ethanol (Figure S12; 2.0 g/kg). We propose that the relative change in firing rate in MD RS neurons might be an important driver and indicator of the change of transition in and out of consciousness as demonstrated for other nuclei of the dMT 8,9,32,48. Therefore, the low variability in firing of MD in Cav3.1 mutant mice might be the driving force for the higher resistance to loss of motion under ethanol. In mutants, brain states might be less distinguishable leading to frequent sleep stages switch23 or resistance to unconsciousness34.

Cav3.1 T-type Calcium and Burst during Low Conscious State

Burst, as a result of Cav3.1 T-type calcium channel de-inactivation/activation, is thought to control the gating of sensory-motor stimuli 57,58 and modulate attention towards novel stimuli rather than the transmission of details 53,57,59. Previous reports highlighted the importance of burst in the stabilization of low level of consciousness 13,22,23 suggesting a direct role for burst, while no mention of the importance of tonic firing was made. We found that the propensity for burst during ethanol-induced LOM (Fig. 4-A1, -B1 and -B2; fLOM: 20/33 bursting neurons; 0.79±1.63 Burst event/min) was lower than in NREM (NREM: 34/34 bursting neurons; 5.76 ± 5.51 burst events/min) and higher than during wakefulness (Wake: 5/36 bursting neurons; 0.16±0.31 Burst event/min). In addition, burst-like electrical stimulation of MD did not significantly affect ethanol resistance (Fig. S10). Although burst-like stimulations are highly artificial and do not recruit T-current and associated mechanisms following low-threshold burst, they allow for the reproduction of the influence of TC burst firing on target centers45, including thalamo-cortical and thalamo- thalamic efferents.

Interestingly, under lower doses of ethanol (I.P. injection of 2.0g/kg of ethanol) mutant and WT alike showed similar levels of resistance to ethanol. We observed a phasic inhibition of MD neurons in WT, capable of inducing partial silencing 60 and rebound bursts 61 (Fig. S12; 1sec ON-OFF using archaerhodopsin-mediated inhibition), did increase fLOM duration mostly (Z(13) = -2.214, p = 0.022, Ranksum test). This result supports that in the context of hypnotic dose of ethanol, the apparition of burst might correlate with unconscious state stability rather than induction. Burst stimulation without inhibition did not have this effect (Fig. S10 and S11). Currently, our data does not allow us to formulate any clear conclusion on the direct role of burst events during fLOM. We propose that the absence of burst and an accompanying effect of maintenance of tonic firing under ethanol in MD was responsible for the observed increase in resistance and maintenance of activity in Cav3.1 mutant mice.

A bidirectional modulation of Cav3.1 expression and alcoholism

In human mutation of Cav3.1 T-type channels exhibit mental disorders including cerebellar ataxia, absence seizure, schizophrenia and autism62. Remarkably, mutation in voltage-gated calcium channels, including Cav3.1 leads to ethanol resistance and alcohol- seeking behavior 21. Reversibly, chronic exposure to ethanol intake is known to impair sleep 63 and increase ethanol resistance. Previous studies have found an alteration in T- type calcium channel expression following chronic exposure to ethanol in non-human primates 64 suggesting that a reduced T-current and the resulting sustained thalamic tonic firing could be a possible mechanism for early stage ethanol resistance in alcoholic subjects, which increases the conversion probability from casual to compulsive consumption of ethanol. The lack of burst and sustained tonic firing might impair the stabilization of sleep, and in turn chronic sleep impairments might engage addiction related networks. Mechanisms such as adenosine receptor depreciation 65,66 or GABA- receptor potentiation 18,67,68 would enhance ethanol resistance and addiction 69,70 spiraling into further sleep fragmentation, memory consolidation deficit, impulsivity and other impairments associated with alcoholism.

Supporting information